---
title: ""
output: html_document
date: ""
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = F, warning = F, eval = F, message = F, cache = T, cache.lazy = T, fig.width = 6, fig.height = 5, fig.align = 'center')

library(here) #v1.0.1
library(dplyr) #v1.0.5
# library(nimble) #v0.12.1
# library(coda)
library(ggplot2)
library(tidyverse)
library(readr)

source(here::here('scripts', 'PlotTheme.R'))

```

```{r vital rates}

library(coda)
out <- readRDS(here::here('results', 'out_sst_probs.RDS'))

all_pars <- colnames(out[[1]])
outmat <- as.matrix(out)

post_sum <- data.frame(
  med = apply(outmat, 2, function(x) quantile(x, probs = 0.5, na.rm = T, names = F)),
  lower = apply(outmat, 2, function(x) quantile(x, probs = 0.025, na.rm = T, names = F)),
  upper = apply(outmat, 2, function(x) quantile(x, probs = 0.975, na.rm = T, names = F)))
post_sum$variable <- row.names(post_sum)

#mean rates
# sex.pars <- all_pars[which(grepl('b.sex', all_pars))]
mu.pars <- all_pars[which(grepl('mu.', all_pars))]

# b.mass_sum <- post_sum %>% filter(grepl('b.mass', variable)) %>%
#   # transform(var_name = ifelse(grepl('b.mass.f', variable), '1',
#   #                                    ifelse(grepl('b.mass.m', variable), '1M',
#                   transform(var_name = ifelse(grepl('b.mass1', variable), '1',
#                        ifelse(grepl('b.massP', variable), 'P', NA))) %>%
#   transform(sex = ifelse(grepl('M', variable), 'M', 'F')) %>%
#   transform(var_name = factor(var_name, levels = c('P', '1')))
#   
# ggplot(b.mass_sum,
#        aes(x = var_name, y = med, col = sex, group = sex)) +
#   geom_errorbar(aes(x = var_name, ymin=lower, ymax=upper),
#                 position = position_dodge(0.5),
#                 width = 0.25, show.legend = F) +
#   geom_point(size = 0.8, position = position_dodge(0.5)) +
#   geom_hline(aes(yintercept = 0), linetype = 'dashed') +
#   xlab('Age class') + ylab(expression(paste('Logit-scale effect of handling mass'))) +
#   ggtitle('') +
#   plot_theme(legend.position = 'top', panel.border = element_rect(fill = NA),
#              plot.title = element_text(hjust = 0.5)) +
#   scale_color_manual(values = rainbow2[c(2,5,6,7,8)], name = 'Sex')


#survival
int_dat <- post_sum %>% filter(grepl('int.phi', variable)) %>%
  # bind_rows(post_sum %>% filter(grepl('int.psi', variable))) %>%
  transform(variable = gsub('int.', '', variable)) %>%
  transform(sex = ifelse(grepl('M', variable), 'M', 'F')) %>%
  transform(age = gsub('phi', '', variable)) %>%
  transform(age = gsub('M', '', age)) %>%
   transform(age = factor(age, levels = c('P', 'PM', '1', 
                                                   '1M', '2.3', '2.3M',
                            '4.6', '4.6M', '7.9', '7.9M', '10.12', '10.12M',
                            '13.16', '13.16M', 'A', 'AM'))) 
  
  
phi_plot <- ggplot(int_dat %>% filter(grepl('phi', variable)),
       aes(age, med, color = sex)) +
  geom_errorbar(aes(x = age, ymin=lower, ymax=upper),
                position = position_dodge(0.5),
                width = 0.35) +
  geom_point(size = 0.8, position = position_dodge(0.5)) +
  xlab('Age class') + ylab(expression(paste('Survival probability ',  '(', phi, ')'))) +
  ggtitle('') +
  # facet_wrap(~sex, scales = 'free_x', drop = T) +
  plot_theme(legend.position = 'top', 
             panel.border = element_rect(fill = NA),
             plot.title = element_text(hjust = 0.5)) +
  scale_color_manual(values = c('black', 'darkgrey'), name = 'Sex')
  # scale_color_manual(values = rainbow2[c(2,6,7,8)], name = 'Sex')

#fecundity
fec_dat <- post_sum %>% filter(grepl('int.fec', variable)) %>%
  # bind_rows(post_sum %>% filter(grepl('int.psi', variable))) %>%
  transform(variable = ifelse(grepl('low', variable), 'Fec_low', 'Fec_high')) %>%
   transform(age = factor(variable, levels = c('Fec_low', 'Fec_high'),
                          labels = c('4:9', '10:22'))) 

fec_plot <- ggplot(fec_dat, aes(x = age, y = value)) +
  geom_errorbar(aes(x = age, ymin=lower, ymax=upper),
                position = position_dodge(0.5),
                width = 0.35) +
  geom_point(size = 0.8, position = position_dodge(0.5)) +
  xlab('Fecundity age bins') + ylab(expression(paste('Fecundity ',  '(', f, ')'))) +
  ggtitle('') +
  # facet_wrap(~sex, scales = 'free_x', drop = T) +
  plot_theme(legend.position = 'top', 
             panel.border = element_rect(fill = NA),
             plot.title = element_text(hjust = 0.5))

#detection
int_dat <- post_sum %>% filter(grepl('int.p', variable)) %>%
  # bind_rows(post_sum %>% filter(grepl('int.psi', variable))) %>%
  transform(variable = gsub('int.', '', variable)) %>%
  filter(!grepl('phi', variable)) %>%
    transform(sex = ifelse(grepl('M', variable), 'M', 'F')) %>%
  transform(age = gsub('p', '', variable)) %>%
  transform(age = gsub('M', '', age)) %>%
    transform(age = factor(age, levels = c('1', '2.3', '4.6',  
                                                   '7.12', 'A'))) 

p_plot <- ggplot(int_dat,
       aes(age, med, color = sex)) +
  geom_errorbar(aes(x = age, ymin=lower, ymax=upper),
                position = position_dodge(0.5),
                width = 0.25) +
  geom_point(size = 0.8, position = position_dodge(0.5)) +
  xlab('Age class') + ylab(expression(paste('Detection probability ',  '(', p, ')'))) +
  ggtitle('') +
  # facet_wrap(~sex, scales = 'free_x', drop = T) +
  plot_theme(legend.position = 'top', 
             panel.border = element_rect(fill = NA),
             plot.title = element_text(hjust = 0.5)) +
  scale_color_manual(values = c('black', 'darkgrey'), name = 'Sex')
  # scale_color_manual(values = rainbow2[c(2,6,7,8)], name = 'Sex')




```

```{r covariate effects and predictions}

sst_raw <- read.csv(here('data', '2025_CCEP_sst.csv'), stringsAsFactors = F, header = T)

#from S. Melin in project summary notes docx
#use pupOcttoJune: pup survival from marking to 1st birthday
#use pupAprtoSept: pup survival via gestational foraging success/pup weight effects
#use nonpupJulytoJune: survival of non-pup age classes - sea lion "year"

sst_dat <- sst_raw %>% filter(time > 1989) #%>% filter(time > 2010)

#sst
b.sst_sum <- post_sum %>% filter(grepl('b.sst', variable)) %>%
  # transform(var_name = ifelse(grepl('b.mass.f', variable), '1', 
  #                                    ifelse(grepl('b.mass.m', variable), '1M',
                  transform(var_name = gsub('b.sst.', '', variable)) %>%
  transform(var_name = factor(var_name, levels = c('P', '1', '2.3', '4.6', '7.9', 'A')))

sst_eff_plot <- ggplot(b.sst_sum, aes(x = var_name, y = med)) +
  geom_errorbar(aes(x = var_name, ymin=lower, ymax=upper),
                position = position_dodge(0.5),
                width = 0.25, show.legend = F) +
  geom_point(size = 0.8, position = position_dodge(0.5)) +
  geom_hline(aes(yintercept = 0), linetype = 'dashed') +
  xlab('Age class') + ylab(expression(paste('Logit-scale effect of SST'))) +
  ggtitle('') +
  pres_theme(legend.position = 'top', #panel.border = element_rect(fill = NA),
             plot.title = element_text(hjust = 0.5))

#changes in survival from sst
## sequence of values
sst_seq <- seq(min(sst_dat[,2]), max(sst_dat[,2]), length.out = 100)
sst_np_seq <- seq(min(sst_dat[,4]), max(sst_dat[,4]), length.out = 100)

# back-transform a single MCMC row across the sequence
get_pred <- function(phi, b_sst, sequence) {
  mu <- plogis(phi)
  logit_p <- mu + b_sst*sequence
  return(1 / (1 + exp(-logit_p)))
}

samples <- as.matrix(out)
#pups
preds_p <- apply(samples, 1, function(s) {
  get_pred(s["int.phiP"], s["b.sst.P"], sst_seq)
})

preds_pm <- apply(samples, 1, function(s) {
  get_pred(s["int.phiPM"], s["b.sst.P"], sst_seq)
})

preds_1 <- apply(samples, 1, function(s) {
  get_pred(s["int.phi1"], s["b.sst.1"], sst_seq)
})

preds_1m <- apply(samples, 1, function(s) {
  get_pred(s["int.phi1M"], s["b.sst.1"], sst_seq)
})

preds_2.3 <- apply(samples, 1, function(s) {
  get_pred(s["int.phi2.3"], s["b.sst.2.3"], sst_seq)
})

preds_2.3m <- apply(samples, 1, function(s) {
  get_pred(s["int.phi2.3M"], s["b.sst.2.3M"], sst_seq)
})

preds_4.6 <- apply(samples, 1, function(s) {
  get_pred(s["int.phi4.6"], s["b.sst.4.6"], sst_seq)
})

# preds_4.6m <- apply(samples, 1, function(s) {
#   get_pred(s["int.phi4.6M"], s["b.sst.4.6M"], sst_seq)
# })

preds_7.9 <- apply(samples, 1, function(s) {
  get_pred(s["int.phi7.9"], s["b.sst.7.9"], sst_seq)
})

# preds_7.9m <- apply(samples, 1, function(s) {
#   get_pred(s["int.phi7.9M"], s["b.sst.7.9M"], sst_seq)
# })

preds_10.12 <- apply(samples, 1, function(s) {
  get_pred(s["int.phi10.12"], s["b.sst.A"], sst_seq)
})

preds_13.16 <- apply(samples, 1, function(s) {
  get_pred(s["int.phi13.16"], s["b.sst.A"], sst_seq)
})

preds_A <- apply(samples, 1, function(s) {
  get_pred(s["int.phiA"], s["b.sst.A"], sst_seq)
})

summarize_preds_pup <- function(pred_matrix) {
  data.frame(
    sst_val = sst_seq,
    val = 1:length(sst_seq),
    mean_phi = rowMeans(pred_matrix),
    low_phi  = apply(pred_matrix, 1, quantile, probs = 0.025),
    high_phi = apply(pred_matrix, 1, quantile, probs = 0.975))
}

summarize_preds_np <- function(pred_matrix) {
  data.frame(
    sst_val = sst_np_seq,
    val = 1:length(sst_np_seq),
    mean_phi = rowMeans(pred_matrix),
    low_phi  = apply(pred_matrix, 1, quantile, probs = 0.025),
    high_phi = apply(pred_matrix, 1, quantile, probs = 0.975))
}

preds_all <- rbind(summarize_preds_pup(preds_p) %>%
                     transform(age = 'P', sex = 'Fem'),
                   summarize_preds_pup(preds_pm) %>%
                     transform(age = 'P', sex = 'Male')) %>%
  bind_rows(rbind(summarize_preds_np(preds_1) %>%
                     transform(age = '1', sex = 'Fem'),
                   summarize_preds_np(preds_1m) %>%
                     transform(age = '1', sex = 'Male'))) %>%
  # bind_rows(rbind(summarize_preds_np(preds_2.3) %>%
  #                    transform(age = '2:3', sex = 'Fem'),
  #                  summarize_preds_np(preds_2.3m) %>%
  #                    transform(age = '2:3', sex = 'Male'))) %>%
  # bind_rows(rbind(summarize_preds_np(preds_4.6) %>%
  #                    transform(age = '4:6', sex = 'Fem'),
  #                  summarize_preds_np(preds_4.6m) %>%
  #                    transform(age = '4:6', sex = 'Male'))) %>%
  bind_rows(summarize_preds_np(preds_7.9) %>%
                     transform(age = '7:9', sex = 'Fem')) %>%
  bind_rows(summarize_preds_np(preds_10.12) %>%
                     transform(age = '10:12', sex = 'Fem')) %>%
  bind_rows(summarize_preds_np(preds_13.16) %>%
                     transform(age = '13:16', sex = 'Fem')) %>%
  bind_rows(summarize_preds_np(preds_A) %>%
                     transform(age = '17+', sex = 'Fem')) %>%
  transform(age = factor(age, levels = c('P', '1', '2:3', '4:6', '7:9',
                                         '10:12', '13:16', '17+')))

#includes both sexes
# sst_preds_plot <- ggplot(preds_all, aes(x = sst_val, y = mean_phi, col = sex)) +
#   geom_ribbon(aes(ymin = low_phi, ymax = high_phi), alpha = 0.2, color = NA) +
#   geom_line(linewidth = 1) +
#   facet_grid(.~age) +
#   labs(x = "SST anomalies",
#        y = "Predicted survival probability") +
#   theme_bw() +
#   theme(legend.position = 'top',
#                 axis.title.x = element_text(margin = margin(t = 7))) +
#   scale_y_continuous(limits = c(0, 1)) + 
#   scale_color_manual(values = rainbow2[c(2,6)], name = 'Sex') + 
#   scale_fill_manual(values = rainbow2[c(2,6)]) #+ 
#   # scale_x_continuous(limits = c(-1.5, 2.5), breaks = c(seq(-1.5,2.5, length = 5)),
#   #                    labels = round(c(seq(0.3,0.65, length = 5)),2))

sst_preds_plot <- ggplot(preds_all %>% filter(sex != 'M'), aes(x = sst_val, y = mean_phi, col = age)) +
  geom_ribbon(data = preds_all, aes(ymin = low_phi, ymax = high_phi, fill = age), 
              alpha = 0.2, color = NA, show.legend = F) +
  geom_line(linewidth = 1) +
  # facet_grid(.~age) +
  labs(x = "SST anomalies",
       y = "Predicted survival probability") +
  theme_bw() +
  theme(legend.position = 'top',
                axis.title.x = element_text(margin = margin(t = 7))) +
  scale_y_continuous(limits = c(0, 1)) + 
  scale_color_manual(values = rainbow2[c(2,4,6,7,8,9)], name = 'Age') + 
  scale_fill_manual(values = rainbow2[c(2,4,6,7,8,9)]) #+ 
  # scale_x_continuous(limits = c(-1.5, 2.5), breaks = c(seq(-1.5,2.5, length = 5)),
  #                    labels = round(c(seq(0.3,0.65, length = 5)),2))

#ENSO "memory" effects
out_ENSO <- readRDS(file = here('results', 'out_ENSO_memory.RDS'))

# all_pars <- colnames(out_ENSO[[1]])
outmat_ENSO <- as.matrix(out_ENSO)

post_sum_ENSO <- data.frame(
  med = apply(outmat_ENSO, 2, function(x) quantile(x, probs = 0.5, na.rm = T, names = F)),
  lower = apply(outmat_ENSO, 2, function(x) quantile(x, probs = 0.025, na.rm = T, names = F)),
  upper = apply(outmat_ENSO, 2, function(x) quantile(x, probs = 0.975, na.rm = T, names = F)))
post_sum_ENSO$variable <- row.names(post_sum_ENSO)


b.enso <- post_sum_ENSO %>% 
  filter(grepl('b.enso', variable)) %>%
  transform(var_name = gsub('b.enso.', '', variable)) %>%
  filter(med != 0) %>%
  transform(var_name = factor(var_name, levels = c('1[1]', '2.3[1]', '4.6[1]', 'A[1]'), 
                              labels = c('1', '2.3', '4.6', '7+')))

enso_eff_plot <- ggplot(b.enso, aes(x = var_name, y = med)) +
  geom_errorbar(aes(x = var_name, ymin=lower, ymax=upper),
                position = position_dodge(0.5),
                width = 0.25, show.legend = F) +
  geom_point(size = 0.8, position = position_dodge(0.5)) +
  geom_hline(aes(yintercept = 0), linetype = 'dashed') +
  xlab('Age class') + ylab(expression(paste('Logit-scale effect of ENSO birth year'))) +
  ggtitle('') +
  pres_theme(legend.position = 'top', #panel.border = element_rect(fill = NA),
             plot.title = element_text(hjust = 0.5))

```


```{r metrics from MR fit}

library(imputeTS)

pup_cnts <- read_tsv(here('data', 'CSL_San Miguel Island_PupCounts_20260622.txt')) %>%
  filter(Year > 1989)

pup_cnt_filled <- na_interpolation(pup_cnts[,1:5]) %>%
  mutate(US_PupCnt = rowSums(across(-Year)))

#stable age logic to initialize N_f and N_m
# define survival rates
n_occasions <- 36
phi_P <- 0.7
phi_1 <- 0.8
phi_J <- 0.85 # Juveniles (Ages 2-5)
phi_A <- 0.92 # Adults (Ages 6-13)
phi_AA <- 0.78

# storage
N_f_track <- matrix(0, nrow = 25, ncol = (n_occasions-1))
N_m_track <- matrix(0, nrow = 25, ncol = (n_occasions-1))

# year 1; assuming 50:50 sex ratio for pups
N_f_track[1, 1] <- round(pup_cnt_filled$US_PupCnt[1] / 2)
N_m_track[1, 1] <- pup_cnt_filled$US_PupCnt[1] - N_f_track[1, 1]

for (a in 1:24) {
  surv <- ifelse(a == 1, phi_P, 
          ifelse(a == 2, phi_1, 
          ifelse(a >= 3 & a <= 5, phi_J,
          ifelse(a >= 6 & a <= 12, phi_A, phi_AA))))
          
  N_f_track[a + 1, 1] <- round(N_f_track[a, 1] * surv)
  N_m_track[a + 1, 1] <- round(N_m_track[a, 1] * surv)
}

# 4. Dynamically project the rest of the matrix forward through time
for (t in 1:(n_occasions - 2)) {
  
  # Insert the REAL pup count for the next year
  N_f_track[1, t+1] <- round(pup_cnt_filled$US_PupCnt[t+1] / 2)
  N_m_track[1, t+1] <- pup_cnt_filled$US_PupCnt[t+1] - N_f_track[1, t+1]
  
  # Pups transition to yearlings (Index 1 -> 2)
  N_f_track[2, t+1] <- round(N_f_track[1, t] * phi_P)
  N_m_track[2, t+1] <- round(N_m_track[1, t] * phi_P)
  
  # Yearlings transition to Age 2 (Index 2 -> 3)
  N_f_track[3, t+1] <- round(N_f_track[2, t] * phi_1)
  N_m_track[3, t+1] <- round(N_m_track[2, t] * phi_1)
  
  # Juveniles transition (Indices 3-5 -> 4-6)
  for (a in 3:5) {
    N_f_track[a+1, t+1] <- round(N_f_track[a, t] * phi_J)
    N_m_track[a+1, t+1] <- round(N_m_track[a, t] * phi_J)
  }
  
  # Adults transition (Indices 6-24 -> 7-25)
  for (a in 6:12) {
    N_f_track[a+1, t+1] <- round(N_f_track[a, t] * phi_A)
    N_m_track[a+1, t+1] <- round(N_m_track[a, t] * phi_A)
  }
  
    for (a in 13:24) {
    N_f_track[a+1, t+1] <- round(N_f_track[a, t] * phi_AA)
    N_m_track[a+1, t+1] <- round(N_m_track[a, t] * phi_AA)
  }
}

# row 1 (pups) is assigned deterministically in the model, NA for inits
N_f_init <- N_f_track
N_m_init <- N_m_track

N_f_init[1, ] <- NA
N_m_init[1, ] <- NA

# use year 1 for for dpois() priors
init_pop_f <- N_f_track[, 1]
init_pop_m <- N_m_track[, 1]


##calculate MNPL and OSP

#simplified approach assuming 60%, or use vital rates from IPM
# MNPL_estimates <- K_estimates * 0.60

#use vital rates from IPM
# extract MCMC matrix 
outmat <- as.matrix(out)

pup_counts <- pup_cnt_filled$US_PupCnt

# Reconstruct abundance deterministically from pup counts and survival
# pup_counts Vector of observed annual pup census counts
# phi_f Matrix of female survival (rows = age classes 1:24, cols = years)
# phi_m Matrix of male survival (rows = age classes 1:24, cols = years)
# init_f Initial female age vector at t=1 (length = max_age)
# init_m Initial male age vector at t=1 (length = max_age)
init_f <- N_f_track[,1]
init_m <- N_m_track[,1]

reconstruct_pop <- function(pup_counts, phi_f, phi_m, init_f, init_m) {
  n_occasions <- length(pup_counts)
  max_age <- nrow(phi_f)
  
  N_f <- matrix(0, nrow = max_age, ncol = n_occasions)
  N_m <- matrix(0, nrow = max_age, ncol = n_occasions)
  
  # Seed Year 1 with starting age distribution
  N_f[,1] <- N_f_track[,1]
  N_m[,1] <- N_m_track[,1]
  
  for (t in 1:(n_occasions - 1)) {
    # pups (Index 1) enter directly from empirical census counts
    # Assumes a 50:50 sex ratio at birth
    N_f[1,t+1] <- pup_counts[t+1]/2
    N_m[1,t+1] <- pup_counts[t+1]/2
    
    # age up older cohorts deterministically using CJS survival rates
    for (a in 1:(max_age - 1)) {
      N_f[a+1,t+1] <- N_f[a,t]*phi_f[a,t]
      N_m[a+1,t+1] <- N_m[a,t]*phi_m[a,t]
    }
  }
  
  # Calculate total abundance per year 
  Ntot <- colSums(N_f) + colSums(N_m)
  
  return(list(Ntot = Ntot, N_f = N_f, N_m = N_m, pups = pup_counts))
}

# Fit Theta-Logistic Model to reconstructed abundance via Sum of Squared Errors
# N_series is a vector of reconstructed total population size over time
fit_theta_logistic <- function(N_series) {
  n_occasions <- length(N_series)
  
  # Objective function: Sum of Squared Errors (SSE)
  sse_fn <- function(params) {
    r     <- params[1]
    K     <- params[2]
    theta <- params[3]
    
    N_pred <- numeric(n_occasions)
    N_pred[1] <- N_series[1]
    
    for (t in 1:(n_occasions - 1)) {
      # Discrete theta-logistic step
      growth <- r * N_pred[t] * (1 - (N_pred[t] / K)^theta)
      N_pred[t+1] <- N_pred[t] + growth
      
      # Penalty for numerical overflow or negative population dips
      if (is.na(N_pred[t+1]) || N_pred[t+1] <= 0) return(1e10)
    }
    
    return(sum((N_series - N_pred)^2))
  }
  
  # Set reasonable biological starting bounds for optimization
  init_K     <- max(N_series) * 1.1
  init_r     <- 0.07 # Standard pinniped intrinsic growth rate
  init_theta <- 2.0  # Standard delayed density dependence
  
  opt <- optim(
    par = c(r = init_r, K = init_K, theta = init_theta),
    fn = sse_fn,
    method = "L-BFGS-B",
    lower = c(r = 0.01, K = max(N_series) * 0.8, theta = 0.5),
    upper = c(r = 0.20, K = max(N_series) * 3.0, theta = 10.0)
  )
  
  # Extract estimates and analytically calculate MNPL
  r_est     <- opt$par["r"]
  K_est     <- opt$par["K"]
  # theta_est <- opt$par["theta"]
  theta_est <- 2.4 #MMPA standard; not always possible to estimate r and K and theta
  mnpl_est  <- K_est * ((1 / (theta_est + 1))^(1 / theta_est))
  
  return(data.frame(
    r = r_est, 
    K = K_est, 
    theta = theta_est, 
    MNPL = mnpl_est, 
    MNPL_ratio = mnpl_est / K_est,
    Convergence = opt$convergence
  ))
}

# loop across CJS posterior draws)

out <- readRDS(here::here('results', 'out_sst_probs.RDS'))
outmat <- as.matrix(out)

nsim <- 500
results_list <- vector("list", nsim)
n_occasions <- 35

phiPcols <- paste0('phiP.prob[', 1:n_occasions, ']')
phi1cols <- paste0('phi1.prob[', 1:n_occasions, ']')
phi2.3cols <- paste0('phi2.3.prob[', 1:n_occasions, ']')
phi4.6cols <- paste0('phi4.6.prob[', 1:n_occasions, ']')
phi7.9cols <- paste0('phi7.9.prob[', 1:n_occasions, ']')
phi10.12cols <- paste0('phi10.12.prob[', 1:n_occasions, ']')
phi13.16cols <- paste0('phi13.16.prob[', 1:n_occasions, ']')
phiAcols <- paste0('phiA.prob[', 1:n_occasions, ']')

phiPMcols <- paste0('phiPM.prob[', 1:n_occasions, ']')
phi1Mcols <- paste0('phi1M.prob[', 1:n_occasions, ']')
phi2.3Mcols <- paste0('phi2.3M.prob[', 1:n_occasions, ']')
phi4.6Mcols <- paste0('phi4.6M.prob[', 1:n_occasions, ']')
phi7.9Mcols <- paste0('phi7.9M.prob[', 1:n_occasions, ']')
phi10.12Mcols <- paste0('phi10.12M.prob[', 1:n_occasions, ']')
phi13.16Mcols <- paste0('phi13.16M.prob[', 1:n_occasions, ']')
phiAMcols <- paste0('phiAM.prob[', 1:n_occasions, ']')

cat("Running post-hoc reconstruction across", nsim, "CJS draws...\n")
for (k in 1:nsim) {

  phiP <- outmat[k,phiPcols]
  phi1 <- outmat[k,phi1cols]
  phi2.3 <- outmat[k,phi2.3cols]
  phi4.6 <- outmat[k,phi4.6cols]
  phi7.9 <- outmat[k,phi7.9cols]
  phi10.12 <- outmat[k,phi10.12cols]
  phi13.16 <- outmat[k,phi13.16cols]
  phiA <- outmat[k,phiAcols]
  
  phiPM <- outmat[k,phiPMcols]
  phi1M <- outmat[k,phi1Mcols]
  phi2.3M <- outmat[k,phi2.3Mcols]
  phi4.6M <- outmat[k,phi4.6Mcols]
  phi7.9M <- outmat[k,phi7.9Mcols]
  phi10.12M <- outmat[k,phi10.12Mcols]
  phi13.16M <- outmat[k,phi13.16Mcols]
  phiAM <- outmat[k,phiAMcols]
  
  phi_f <- rbind(phiP, phi1, phi2.3, phi2.3, phi4.6, phi4.6, phi4.6, 
                 phi7.9, phi7.9, phi7.9, phi10.12, phi10.12, phi10.12, 
                 phi13.16, phi13.16, phi13.16, phi13.16,
                 phiA, phiA, phiA, phiA, phiA, phiA, phiA, phiA)
  phi_m <- rbind(phiPM, phi1M, #phi2.3M, phi2.3M, phi4.6M, phi4.6M, phi4.6M,
                 phi2.3, phi2.3, phi4.6, phi4.6, phi4.6,
                 phi7.9M, phi7.9M, phi7.9M, phi10.12M, phi10.12M, phi10.12M, 
                 phi13.16M, phi13.16M, phi13.16M, phi13.16M, 
                 phiAM, phiAM, phiAM, phiAM, phiAM, phiAM, phiAM, phiAM)

  # Run deterministic projection
  recon <- reconstruct_pop(pup_counts, phi_f, phi_m, init_f, init_m)
  
  # Fit theta-logistic model
  fit <- fit_theta_logistic(recon$Ntot)
  fit$draw <- k
  results_list[[k]] <- fit
}

# Combine results into a final summary dataframe
post <- bind_rows(results_list)

# View summary statistics 
summary_table <- post %>%
  summarise(
    K_median    = median(K),
    K_LCI       = quantile(K, 0.025),
    K_UCI       = quantile(K, 0.975),
    MNPL_median = median(MNPL),
    MNPL_LCI    = quantile(MNPL, 0.025),
    MNPL_UCI    = quantile(MNPL, 0.975),
    Theta_median = median(theta),
    Theta_LCI = quantile(theta, 0.025),
    theta_UCI = quantile(theta, 0.975)
  )

# print(summary_table)


# calculate the implied fecundity
# How many pups were born per adult female (assuming ages 4-24 are adults)?
adult_females <- colSums(recon$N_f[4:24, ])
implied_fecundity <- pup_counts / adult_females

# Ntot_recon_plot <- plot(x = 1990:2024, y = recon$Ntot, type = "l", col = "darkblue", lwd = 2, 
#      main = "Reconstructed population trajectory",
#      ylab = "Total abundance", xlab = "")
# 
# fec_calc_plot <- plot(x = 1990:2024, y = implied_fecundity, type = "b", col = "darkred", pch = 16,
#      # main = "Implied birth rate",
#      ylab = "Fecundity (pups per adult female)", xlab = "")


```

```{r metrics from IPM fit and post pred checks}

out_IPM <- readRDS(file = here('results', 'out_IPM_sst_DD_slice2.RDS'))
outmat <- do.call(rbind, out_IPM)

outmat <- as.matrix(out_IPM)

post_sumIPM <- data.frame(
  med = apply(outmat, 2, function(x) quantile(x, probs = 0.5, na.rm = T, names = F)),
  lower = apply(outmat, 2, function(x) quantile(x, probs = 0.025, na.rm = T, names = F)),
  upper = apply(outmat, 2, function(x) quantile(x, probs = 0.975, na.rm = T, names = F)))
post_sumIPM$variable <- row.names(post_sumIPM)

#survival
int_datIPM <- post_sumIPM %>% filter(grepl('int.phi', variable)) %>%
  # bind_rows(post_sum %>% filter(grepl('int.psi', variable))) %>%
  transform(variable = gsub('int.', '', variable)) %>%
  transform(sex = ifelse(grepl('M', variable), 'M', 'F')) %>%
  transform(age = gsub('phi', '', variable)) %>%
  transform(age = gsub('M', '', age)) %>%
   transform(age = factor(age, levels = c('P', 'PM', '1', 
                                                   '1M', '2.3', '2.3M',
                            '4.6', '4.6M', '7.9', '7.9M', '10.12', '10.12M',
                            '13.16', '13.16M', 'A', 'AM'))) %>%
  transform(mod = 'IPM') %>%
  bind_rows(int_dat %>% transform(mod = 'MR'))
  
plot_dat <- int_datIPM %>% filter(sex != 'M')

phi_plot_compare <- ggplot(plot_dat, aes(age, med, color = mod)) +
  geom_errorbar(aes(x = age, ymin=lower, ymax=upper),
                position = position_dodge(0.5),
                width = 0.35) +
  geom_point(size = 0.8, position = position_dodge(0.5)) +
  xlab('Age class') + ylab(expression(paste('Survival probability ',  '(', phi, ')'))) +
  ggtitle('') +
  # facet_wrap(~sex, scales = 'free_x', drop = T) +
  plot_theme(legend.position = 'top', 
             panel.border = element_rect(fill = NA),
             plot.title = element_text(hjust = 0.5)) +
  scale_color_manual(values = c('black', 'darkgrey'), name = 'Model')

#fecundity
fec_dat <- data.frame(outmat) %>% dplyr::select(contains('int.fec'))
colnames(fec_dat) <- c('Age 10+', 'Age 4-9')

fec_plot <- ggplot(fec_dat %>% reshape2::melt() %>%
                     transform(variable = factor(variable, levels = c('Age 4-9', 'Age 10+'))), 
                   aes(x = variable, y = value)) +
  geom_violin() +
  ylab('Fecundity') xlab('') +
  theme_bw()

DD_sum <- post_sumIPM %>% filter(grepl('beta.DD', variable))

DD_plot <- ggplot(DD_sum, aes(y = med, x = variable)) +
  geom_point(aes(x = variable, y = med)) +
  geom_errorbar(aes(x = variable, ymin = lower, ymax = upper), width = 0.1) +
  ylab('Density dependence') + xlab('') +
  theme_bw()

##PPC
# expected pups columns chronologically
pup_cols <- paste0("expected_pups_tot[", 2:n_occasions, "]") # Starts at 2 because t+1 in the loop
pup_post <- outmat[, pup_cols]

# calculate the median and 95% CI for each year and append census data 
pup_summary <- data.frame(
  TimeStep = 2:n_occasions, 
  year = 1991:2024,
  Expected_Median = apply(pup_post, 2, median),
  Expected_LCI = apply(pup_post, 2, quantile, probs = 0.025),
  Expected_UCI = apply(pup_post, 2, quantile, probs = 0.975),
  Observed = pup_cnt_filled$US_PupCnt[2:n_occasions]
)

pups_ppc_plot <- ggplot(pup_summary, aes(x = year)) +
  geom_ribbon(aes(ymin = Expected_LCI, ymax = Expected_UCI), 
              fill = "steelblue", alpha = 0.3) +
  geom_line(aes(y = Expected_Median), color = "steelblue", size = 1) +
  geom_point(aes(y = Observed), color = "black", size = 2) +
  geom_line(aes(y = Observed), linetype = 'dashed', col = 'darkgrey') +
  theme_bw() +
  labs(
    title = "Posterior predictive check",
    # subtitle = "Black dots = Census Data | Blue line & ribbon = Model Expected Pups (95% CI)",
    x = "",
    y = "Total pups") +
  plot_theme()


#### also plot scatterplot of int.phiP and beta.DD?






##calculate K and MNPL
# Number of posterior draws to simulate (e.g., sample 1000 to save time)
n_draws <- 500#0
draw_indices <- sample(1:nrow(outmat), n_draws)

# Vector to store K estimates
K_estimates <- numeric(n_draws)

# Run the simulation for each sampled posterior draw
for (i in 1:n_draws) {
  row_idx <- draw_indices[i]

  # extract vital rates for each MCMC draw
  int_phiP   <- outmat[row_idx, "int.phiP"]
  beta_DD <- outmat[row_idx, "beta.DD"]
  int_phi1   <- outmat[row_idx, "int.phi1"]
  int_phi2.3   <- outmat[row_idx, "int.phi2.3"]
  int_phi4.6   <- outmat[row_idx, "int.phi4.6"]
  int_phi7.9   <- outmat[row_idx, "int.phi7.9"]
  int_phi10.12   <- outmat[row_idx, "int.phi10.12"]
  int_phi13.16   <- outmat[row_idx, "int.phi13.16"]
  int_phiA   <- outmat[row_idx, "int.phiA"]
  fec_low  <- outmat[row_idx, "int.fec_low"]
  fec_high <- outmat[row_idx, "int.fec_high"]

  # convert baseline survival to probability scale (assuming SST = 0)
  mu_P <- qlogis(int_phiP)
  mu_1 <- qlogis(int_phi1)
  mu_2.3 <- qlogis(int_phi2.3)
  mu_4.6 <- qlogis(int_phi4.6)
  mu_7.9 <- qlogis(int_phi7.9)
  mu_10.12 <- qlogis(int_phi10.12)
  mu_13.16 <- qlogis(int_phi13.16)
  mu_A <- qlogis(int_phiA)

  # initialize a starting population vector
  # (e.g., 50 females per age class to the simulation)
  N_vec <- rep(50, 25)

  # project forward to reach equilibrium (K)
  for (y in 1:500) { # 200 years long enough?

    N_tot <- sum(N_vec) # just modeling females here for simplicity,
                        # or multiply by 2 if assuming equal male population.

    # calculate density-dependent pup survival for THIS year
    # scale N as in the model
    phiP_current <- plogis(mu_P-beta_DD*((N_tot*2)/100000))

    # calculate total pups born
    pups_born <- (sum(N_vec[5:10])*fec_low*0.5) + (sum(N_vec[11:23])*fec_high*0.5)

    # Create the next year's population vector
    N_vec_next <- numeric(25)
    N_vec_next[1] <- pups_born
    N_vec_next[2] <- N_vec[1] * phiP_current
    N_vec_next[3] <- N_vec[2] * int_phi1
    N_vec_next[4] <- N_vec[3] * int_phi2.3
    N_vec_next[5] <- N_vec[4] * int_phi2.3
    N_vec_next[6] <- N_vec[5] * int_phi4.6
    N_vec_next[7] <- N_vec[6] * int_phi4.6
    N_vec_next[8] <- N_vec[7] * int_phi4.6
    N_vec_next[9] <- N_vec[8] * int_phi7.9
    N_vec_next[10] <- N_vec[9] * int_phi7.9
    N_vec_next[11] <- N_vec[10] * int_phi7.9
    N_vec_next[12] <- N_vec[11] * int_phi10.12
    N_vec_next[13] <- N_vec[12] * int_phi10.12
    N_vec_next[14] <- N_vec[13] * int_phi10.12
    N_vec_next[15] <- N_vec[14] * int_phi13.16
    N_vec_next[16] <- N_vec[15] * int_phi13.16
    N_vec_next[17] <- N_vec[16] * int_phi13.16
    N_vec_next[18] <- N_vec[17] * int_phi13.16

    # absorbing state (ages 17-23)
    for (a in 18:24) {
      N_vec_next[a+1] <- N_vec[a] * int_phiA
    }

    # update population
    N_vec <- N_vec_next
  }

  # store stabilized population size (multiply by 2 to account for males)
  K_estimates[i] <- sum(N_vec) * 2
}

# mean
mean_K <- mean(K_estimates)
K_CI <- quantile(K_estimates, probs = c(0.025, 0.975))

# print(paste("Estimated K:", round(mean_K), "95% CI:", round(K_CI[1]), "-", round(K_CI[2])))

#net productivity curve
####derive net productivity curve

# extract your posterior medians
# Baseline survival (converted to logit scale for the density penalty math)
mu_P_f <- qlogis(median(outmat[, "int.phiP"]))
mu_P_m <- qlogis(median(outmat[, "int.phiPM"]))

# Define a helper vector for female and male adult survival across indices 2 to 24
# You will need to map your specific age-class medians (int.phi1, int.phi2.3, etc.) 
# to the 23 transitions (Index 1 to 2, up to Index 24 to 25).
# For this example, I am using a simplified constant adult survival for brevity:
phi_adult_f <- rep(median(outmat[, "int.phiA"]), 24) 
phi_adult_m <- rep(median(outmat[, "int.phiAM"]), 24) 

# Fecundity and Density Dependence
fec_low <- median(outmat[, "int.fec_low"])
fec_high <- median(outmat[, "int.fec_high"])
beta_DD <- median(outmat[, "beta.DD"])

# 2. Setup the projection
n_years <- 300
N_f <- matrix(0, nrow = 25, ncol = n_years)
N_m <- matrix(0, nrow = 25, ncol = n_years)

# Seed a small founder population at t = 1
N_f[, 1] <- 100
N_m[, 1] <- 100

total_N <- numeric(n_years)
delta_N <- numeric(n_years)

total_N[1] <- sum(N_f[, 1]) + sum(N_m[, 1])

# 3. Project the population forward deterministically
for (t in 1:(n_years - 1)) {
  
  # Calculate density-dependent pup survival for this year
  # (SST is assumed to be 0 for an average environment)
  N_scaled <- total_N[t] / 100000
  phi_P_f_current <- plogis(mu_P_f - (beta_DD * N_scaled))
  phi_P_m_current <- plogis(mu_P_m - (beta_DD * N_scaled))
  
  # A. Transition the survivors first (Index 1 to 24 surviving to 2 to 25)
  N_f[2, t+1] <- N_f[1, t] * phi_P_f_current
  N_m[2, t+1] <- N_m[1, t] * phi_P_m_current
  
  for(a in 2:24) {
    N_f[a+1, t+1] <- N_f[a, t] * phi_adult_f[a]
    N_m[a+1, t+1] <- N_m[a, t] * phi_adult_m[a]
  }
  
  # B. Calculate post-breeding reproduction based on the new adult pool
  pups_tot <- (sum(N_f[5:10, t+1]) * fec_low) + (sum(N_f[11:23, t+1]) * fec_high)
  
  # Assign pups to Index 1 (50:50 sex ratio)
  N_f[1, t+1] <- pups_tot / 2
  N_m[1, t+1] <- pups_tot / 2
  
  # C. Record population metrics
  total_N[t+1] <- sum(N_f[, t+1]) + sum(N_m[, t+1])
  delta_N[t] <- total_N[t+1] - total_N[t]
}

# The final delta_N is NA because we don't project to t+1 at the end
delta_N[n_years] <- 0 

# 4. Find K and MNPL
# K is the maximum population size reached (where delta_N hits 0)
K_est <- max(total_N)
# MNPL is the population size where growth (delta_N) is maximized
MNPL_est <- total_N[which.max(delta_N)]

# 5. Plot the Net Productivity Curve
prod_data <- data.frame(N = total_N, Growth = delta_N) %>%
  filter(Growth > 0) # Remove the zero-growth equilibrium tail for a cleaner plot

ggplot(prod_data, aes(x = N, y = Growth)) +
  geom_line(color = "darkgreen", size = 1.2) +
  
  # Vertical line for MNPL
  geom_vline(xintercept = MNPL_est, linetype = "dashed", color = "blue") +
  annotate("text", x = MNPL_est, y = max(prod_data$Growth) * 0.9, 
           label = paste("MNPL =", round(MNPL_est)), color = "blue", hjust = -0.1) +
  
  # Vertical line for K
  geom_vline(xintercept = K_est, linetype = "dashed", color = "red") +
  annotate("text", x = K_est, y = max(prod_data$Growth) * 0.5, 
           label = paste("K =", round(K_est)), color = "red", hjust = 1.1) +
  
  theme_minimal() +
  labs(
    title = "Emergent Net Productivity Curve",
    x = "Total Population Size (N)",
    y = "Net Annual Growth (\u0394 N)"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.title = element_text(face = "bold")
  )

# the curve shouldn't be a perfectly symmetrical parabola. It should rise steadily, peak (the blue MNPL line) noticeably to the right of the halfway mark ($0.5K$), and then crash steeply down to zero at carrying capacity (the red $K$ line). That right-skew is the biological signature of delayed density dependence acting exclusively on pup survival.

```


```{r temporal variance}

phi.vals <- post_sum %>%
  filter(grepl('.prob', variable)) %>%
    transform(year = as.numeric(as.character(gsub('.+\\[([0-9]+)\\].*$', '\\1',
                                                variable)))) %>%
  #use when dataframe for phi.prob rather than vector
  # transform(year = as.numeric(as.character(gsub('.+\\[([0-9]+), ([0-9]+)\\].*$', '\\2', 
  #                                               variable)))) %>% 
  # transform(region = gsub('.+\\[([0-9]+), ([0-9]+)\\].*$', '\\1', variable)) %>%
  transform(variable = sub("\\[.*", "", variable)) %>%
  transform(rate = ifelse(grepl('phi', variable), 'Survival', 'Detection')) %>%
  transform(age = gsub('.prob', '', variable)) %>%
  transform(age = ifelse(rate == 'Survival', gsub('phi', '', age), 
                         ifelse(rate == 'Detection', gsub('p', '', age), NA))) %>%
  transform(sex = ifelse(grepl('M', age), 'Male', 'Female')) %>%
  transform(age = gsub('M', '', age)) %>%
  transform(age = factor(age, levels = c('P', '1', '2.3', '4.6', '7.9', '10.12', '13.16', '7.12', 'A'))) %>%
  # merge(mean.phi.vals %>% filter(site == 'Pyramid Rock'), by = c('age'), suffixes = c('', '.mean'))
  transform(exclude_det = ifelse(age == '2.3' & year < 2, 'Y',
                                   ifelse(age == '4.6' & year < 4, 'Y',
                                    ifelse(age == '7.12' & year < 7, 'Y',
                                           ifelse(age == 'A' & year < 13, 'Y', 'N'))))) %>%
  transform(exclude_phi = ifelse(age == '1' & year < 2, 'Y',
                             ifelse(age == '2.3' & year < 3, 'Y',
                                   ifelse(age == '4.6' & year < 5, 'Y',
                                    ifelse(age == '7.9' & year < 7, 'Y',
                                           ifelse(age == '10.12' & year < 10, 'Y',
                                                  ifelse(age == '13.16' & year < 13, 'Y',
                                           ifelse(age == 'A' & year < 16, 'Y', 'N'))))))))

#temporal variance with mean rates; intercept represents female, so can't show males
ggplot(phi.vals %>% filter(rate == 'Detection' & exclude_det != 'Y' & year != 20),
       aes(year, med)) +
  geom_errorbar(aes(x = year, ymin=lower, ymax=upper), position = position_dodge(0.5),
                width = 0.5, show.legend = F) +
  geom_point(size = 0.8, position = position_dodge(0.5)) +
  geom_line(linewidth = 0.7, position = position_dodge(0.5)) +
  # geom_line(aes(year, med.mean), col = 'grey20', linetype = 'dotted') +
  xlab('') + ylab(expression(paste('Detection probability ',  '(', p, ')'))) +
  ggtitle('') +
    facet_grid(age ~ sex) +
  plot_theme(legend.position = 'top', panel.border = element_rect(fill = NA),
             plot.title = element_text(hjust = 0.5)) +
  guides(color = guide_legend("", nrow = 1, byrow = T)) +
  scale_color_manual(values = rainbow2[c(2,3,5,6,7,8)], name = '') +
  scale_x_continuous(breaks = c(seq(1, 35, by = 5)), 
                     labels = c(seq(1990, (1990+n_occasions-1), by = 5)))

#survival
ggplot(phi.vals %>% filter(rate == 'Survival' & exclude_phi != 'Y'),
       aes(year, med)) +
  geom_errorbar(aes(x = year, ymin=lower, ymax=upper), position = position_dodge(0.5),
                width = 0.5, show.legend = F) +
  geom_point(size = 0.8, position = position_dodge(0.5)) +
  geom_line(linewidth = 0.7, position = position_dodge(0.5)) +
  # geom_line(aes(year, med.mean), col = 'grey20', linetype = 'dotted') +
  xlab('') + ylab(expression(paste('Survival probability ',  '(', phi, ')'))) +
  ggtitle('') +
    facet_grid(age ~ sex) +
  plot_theme(legend.position = 'top', panel.border = element_rect(fill = NA),
             plot.title = element_text(hjust = 0.5)) +
  guides(color = guide_legend("", nrow = 1, byrow = T)) +
  scale_color_manual(values = rainbow2[c(2,3,5,6,7,8)], name = '') +
  scale_x_continuous(breaks = c(seq(1, 25, by = 5)), labels = c(seq(2000, (2000+n_occasions-1), by = 5)))

ggplot(phi.vals %>% filter(rate == 'Survival' & exclude_phi != 'Y' &
                             age %in% c('P', '1') & sex == 'Female'),
       aes(year, med)) +
  geom_errorbar(aes(x = year, ymin=lower, ymax=upper), position = position_dodge(0.5),
                width = 0.5, show.legend = F) +
  geom_point(size = 0.8, position = position_dodge(0.5)) +
  geom_line(linewidth = 0.7, position = position_dodge(0.5)) +
  # geom_line(aes(year, med.mean), col = 'grey20', linetype = 'dotted') +
  xlab('') + ylab(expression(paste('Survival probability ',  ' (', phi, ')'))) +
  ggtitle('') +
    facet_grid(age ~ sex) +
  # theme_classic() +
  pres_theme(legend.position = 'top', #panel.border = element_rect(fill = NA),
             plot.title = element_text(hjust = 0.5)) +
  guides(color = guide_legend("", nrow = 1, byrow = T)) +
  scale_color_manual(values = rainbow2[c(2,3,5,6,7,8)], name = '') +
  scale_x_continuous(breaks = c(seq(1, 25, by = 5)), labels = c(seq(2000, (2000+n_occasions-1), by = 5)))


```

#### CSL population dynamics and environmental effects  

*Project status*  

- MR model estimates look great  
-- Haven't incorporated dead stranded IDs or lepto/DA  
- Prototype IPM model estimates look like they're working (model takes a *long* long time to run)  
- Estimates of management metrics between IPM and MR models align - this is also great and not to be taken for granted   
- Methods drafted, results visualized  
- Medium to-do.... think a bit about the density dependence mechanism in the IPM given the estimated effects of ENSO on not just pups... yearlings and juveniles have a harder time finding food too? 

*Draft fodder for future manuscript effort...* 

#### Introduction 

CSL as indicator species in a dynamic CCE 
- Importance of monitoring sentinel species to understand population dynamics, provide updated management and conservation metrics, and understand responses of a species sensitive to environmental variability in an ecosystem notorious for cyclic phases  
- Climate impacts the food web and health  
- Opportunities afforded by a well-studied (and observable) species in a dynamic environment  

Estimating density dependence for long-lived top predator  
- Challenges  
- Different approaches  

CSL historical and contemporary context  
- ups and downs and connections with ENSO  
- prominent species along the coast  
- foraging ecology/fisheries interactions  
- human impacts and disease/stranding/public concern  

Current importance  
-- estimates growing outdated  
-- improvements upon previous estimation, and more SST evidence  

Aims: upgrade Laake et al. and provide management metrics    
- Laake et al implemented a two-step approach: survival estimates fed into population projection model and then separately derive K and MNPL  
-- Benefits: computationally efficient  
-- Disadvantages: (1) does not fully integrate uncertainty in MR estimates within the population projection model and therefore does not resolve discrepancies between the two datasets based on relative precision of the data streams; (2) DD not estimated with respect to a particular vital rate (doesn't specify mechanism of dependence) and assumes downturns in abundance are due to density instead of environmental effect  
- Fully integrated IPM addresses these issues: (1) weights uncertainty in each data stream to settle discrepancies; uncertainty from all components propagated into final posteriors of *K* and MNPL, and (2) estimates DD with respect to an ecological mechanism and facilitates specifying how transient dynamics (such as small cohort of pups from ENSO event) propagates through the age structure over the next decade   
- Full IPM distinguishes between ENSO noise and density signal - what would *K* be in an average environment -- may be a higher/truer carrying capacity instead of logistic growth model where a lot of high SST years could look like approaching *K*   
- In the IPM, MNPL and *K* are emergent properties derived from age-structured vital rates. Laake et al. used the $\theta$ parameter to force the productivity curve into a specific mathematical shape, which could skew MNPL relative to *K* (generally 0.6 for marine mammals instead of 0.5K that would be represented by perfect symmetry) whereas the IPM will organically generate right-skewed productivity curve based on vital rates  
- Ultimately, this model enhances ecological realism  

<!-- Literature: -->
<!-- - Hammill et al. 2021 for northwest atlantic grey seals; similar density dependence framework -->
<!-- - Ersalman et al. 2024-2025 on baltic ringed seals? env focus on younglings -->
<!-- - Tinker et al. 2023 for NW atlantic harp seals; K and ice-related pup mortality -->

<!-- Hammill et al. and Tinker et al still use mathematical shape parameter theta -->
 
#### Methods

##### Study system   

##### Data collection and curation  

- Marked cohort sizes and years (every year? similar approximate samples sizes?)
-- approximately 500 marked annually (!!wow!!) 1990-2005, 300-400 from 2006-2015, 250-300 thereafter.

<!-- (table(ch_dat_m$cohort) + table(ch_dat_f$cohort)) -->

- Pup counts and imputation to fill any gaps.

- SST data and seasonality as relevant to pups vs. non-pups.


##### Statistical analyses

We developed a fully integrated Bayesian population model to jointly estimate vital rates and population dynamics for California sea lions using data collected from the California Coast from 1990-2025. The IPM framework simultaneously evaluates a Cormack-Jolly-Seber (CJS; citation) mark-recapture model and a state-space demographic projection matrix, linked by shared age- and sex-specific survival parameters. To resolve latent fecundity and anchor population trajectories, we incorporated annual pup counts treated as a census without observation error.

*Vital rates subcomponent model*

We modeled the individual capture histories using a CJS framework to estimate age- and sex-specific apparent survival ($\phi$) and detection probabilities ($p$). The state process was modeled using a categorical distribution tracking individuals across distinct biological age classes, up to a maximum modeled lifespan of 24 years.

Survival probabilities were estimated for pups, ages 1, 2-3, 4-6, 7-9, 10-12, 13-16, and 17+ for both males and females. Survival probabilities were estimated on the logit scale and modeled as a function of baseline age- and sex-specific intercepts ($\mu$) and environmental stochasticity ($\beta_{\text{SST}}$). Additional temporal variation not attributable to the effects of SST were not included and pup birth weight, while known to be influential in survival outcomes, was not included as a covariate here due to its correlation with annual SST (i.e., pup weights negatively correlated with SST; citation or create additional supplemental figure) . Because density dependence in pinnipeds acts most strongly on early life stages (i.e., pups), so the density effect ($\beta_{\text{DD}}$) was isolated to pup survival ($\phi_P$). Total annual population size, $N_t$, was the summation of all male and female age classes and scaled by a constant ($100,000$) to facilitate Markov chain Monte Carlo (MCMC) convergence. For an individual $i$ at time $t$ and sex $s$, pup survival was parameterized as:

$$\text{logit}(\phi_{p, i, t}) = \mu_p + \beta_{\text{SST}_p} \cdot \text{SST}_{t} - \beta_{\text{DD}} \cdot \left( \frac{N_{t}}{100,000} \right)$$

Juvenile and adult survival and fecundity were assumed to be density-independent and driven solely by baseline rates and environmental stochasticity. Thus, survival for older age classes (yearlings through adults) was parameterized with age-specific intercepts and distinct sea surface temperature values ($\text{SST}_{t}$) relevant to foraging ecology:

$$\text{logit}(\phi_{a, i, t}) = \mu_a + \beta_{\text{SST}, a} \cdot \text{SST}_{t}$$
Detection probailities ($p$) were assumed to be constant between sexes and estimated for each age class with temporal variance pooled across ages estimated using a shrinkage prior, but detection was assumed to be constant across sexes, natal regions, and resight location. 

In the mark-resight subcomponent model, we assumed that marks were not lost, misread, or missed when an animal is encountered, and that all individuals within a specific age and sex class share the same underlying probabilities of survival and detection in a given year.

*Population projection subcomponent model*

To evaluate population-level dynamics, we constructed a sex-specific deterministic and stochastic projection model using survival rates estimated as above. We assumed a post-breeding census timeline, meaning the annual pup counts represent reproduction by females that successfully survived from year $t$ into year $t+1$.

Because pup counts were treated as a census, the number of pups at age 0 ($N_{1, t}$) was assigned deterministically from field observations, assuming a 50:50 sex ratio. To avoid MCMC initialization failures, the latent age classes for the first year of the time series were initialized using a stable age distribution derived from baseline survival probability approximations.

Population structure and age-specific abundance for each age class $a$ were modeled via Poisson distributions (to facilitate MCMC sampling instead of more constrained binomial draws) using the predicted annual age-specific survival probabilities (${\phi}_{a, t}$) accounting for changes in SST over the study period multiplied by the age-specific abundance in the previous year:

$$N_{a+1, t+1} \sim \text{Poisson}({\phi}_{a, t}\cdot N_{a, t})$$

Individuals reaching the terminal age class (age 24) were assumed to undergo complete reproductive and survival senescence and were removed from the matrix in subsequent time steps.

*Integrated population model*

To link the unobserved adult female population to the perfectly observed pup census, we modeled reproduction using a post-breeding census assumption. Pups born in year $t+1$ were calculated as a function of the females that successfully survived into year $t+1$.

Fecundity was estimated as a latent parameter divided into two biological tiers: a lower fecundity rate for younger reproductive females ($F_{\text{4:9}}$, Ages 4–9) and a higher fecundity rate for prime-age adults ($F_{\text{10:22}}$, Ages 10–22). Assuming an equal sex ratio at birth, the expected total number of pups produced was calculated as:

$$E[Pups_{t+1}] = \left( \sum_{a=5}^{10} N_{f, a, t+1} \cdot F_{\text{4:9}} \right) + \left( \sum_{a=11}^{23} N_{f, a, t+1} \cdot F_{\text{10:22}} \right)$$

The final integration of the demography and the count data was achieved by evaluating the expected pup production as estimated in the population growth equations in terms of the census data ($y_{t+1}$) using a Poisson likelihood constraint:

$$y_{t+1} \sim \text{Poisson}(E[Pups_{t+1}])$$

Any divergence between expected reproduction and the census data directly penalizes the likelihood of the adult female survival and fecundity estimates, ensuring mathematical tension and precise demographic alignment across the integrated framework.

Assuming the mark-recapture data ($\mathbf{y}_{MR}$) and the pup census counts ($\mathbf{y}_{count}$) are conditionally independent, the joint likelihood of the model ($L_{IPM}$) is the product of their separate likelihoods:

$$L_{IPM}(\theta|\mathbf{y}_{MR},\mathbf{y}_{count})=L_{MR}(\theta|\mathbf{y}_{MR})\times L_{count}(\theta|\mathbf{y}_{count})$$
where $\theta$ represents the joint vector of all estimated parameters (survival, detection, fecundity, and environmental/density coefficients). By evaluating this joint likelihood, the model allows the high-precision pup census data to inform and constrain the latent vital rates that cannot be fully resolved by the mark-recapture data alone.

*Derivation of carrying capacity ($K$)*
Because density dependence was parameterized mechanistically as a penalty on pup survival rather than through an overarching population growth curve, carrying capacity ($K$) was not estimated as an explicit parameter in the joint likelihood. Instead, $K$ was treated as an emergent property of the estimated vital rates.For each posterior draw, $K$ was derived numerically via a post-hoc simulation. The estimated age-specific survival and fecundity rates were used to project a theoretical population forward in time under average environmental conditions ($\text{SST}_{t}=0$). As the simulated population grew, the density-dependent penalty ($\beta_{DD}$) incrementally suppressed pup survival. $K$ was defined as the equilibrium population size at which the number of deaths perfectly balanced the number of births (population growth rate $\lambda=1$):

$$K=\lim_{t\to\infty}\sum_{a=1}^{25}N_{a,t}\quad\text{where}\quad N_{t+1}=N_{t}$$

*Estimation of MNPL and OSP Status*  

To evaluate the population status relative to the Marine Mammal Protection Act (MMPA) management metrics, we calculated the Maximum Net Productivity Level (MNPL) and Optimal Sustainable Population (OSP) status from the posterior draws of the $K$ simulations. 

Rather than assuming a rigid generalized logistic (theta-logistic) shape parameter, MNPL was derived numerically by tracking the absolute population growth ($\Delta N$) at each time step of the *post-hoc* simulation. MNPL was identified as the exact population size where net production (the number of individuals added to the population annually) reached its global maximum:

$$MNPL=N_{t}\quad\text{where}\quad\max(N_{t+1}-N_{t})\text{ occurs}$$

Finally, the OSP status was evaluated as the ratio of the most recent observed total population size ($N_{current}$) to the derived MNPL. A population is considered to be within its OSP if this ratio is greater than or equal to 1:

$$OSP_{status}=\frac{N_{current}}{MNPL}$$

This numeric derivation ensures that the management metrics are driven entirely by the species' empirical, age-structured vital rates and specific density-dependent survival thresholds estimated by the IPM.

##### Implementation and diagnostics 

This Bayesian state-space integrated population model was implemented in NIMBLE (NIMBLE Development Team 2019) using three Markov chain Monte Carlo (MCMC) chains with 30,000 iterations, 10,000 burn-in, and a thinning rate of 3 within the R programming environment (R Core Team 2020). Convergence was assessed by examining MCMC chains and a Brooks-Gelman-Rubin statistic (Gelman & Rubin 1992; Brooks & Roberts 1998) < 1.1. We summarized the results of the full model according to the proportion of MCMC chain samples for the environmental effects that was above versus below zero. 

#### Results  

Results described from the IPM are still only a subset (50-100 marked individuals per year) of the full dataset. 

##### Vital rates  

```{r phi plot}

phi_plot
phi_plot_compare

```

The first plot shows sex- and age class-specific survival probabilities and associated 95% credible intervals. The second plot shows very strong agreement between these MR estimates and those from the IPM for female survival probabilities. 


```{r p plot}

p_plot

```

Sex- and age-specific annual detection probabilities, though yearlings seem low? Should it be sex-specific for yearlings and juveniles too? Males lower makes sense too?  

##### Covariate effects  

```{r cov plots}

sst_eff_plot
sst_preds_plot
enso_eff_plot

```

Figure (A): logit-scale (positive versus negative) effect of SST for each age shows that pups and yearlings are largely affected similarly and that there are negative effects estimated for all age classes except 4-6yo's. Plain language interpretation: survival is lower for all age classes (except 4-6) in years when SST is high and the effect is strongest for pups and yearlings. *Note to self - recheck whether these are sex-specific effects?*

Figure (B): probability-scale effects of the SST covariate effect for each age: survival goes from approximately 75% in years with lower SST to approximately 30% for pups and yearlings when SST is high. 

Figure (C): I'm calling this the "ENSO memory" effect: whether individuals are affected in subsequent years if born in an ENSO year. This shows that (for example) survival is high at ages 2-6 for individuals born in ENSO years and lower for adults that were born in ENSO years. We should talk through this a bit more, but my guess is that this is more showing us a cyclic effect rather than some sort of lasting ENSO effect that pups carry with them as they age. Once pups that were born in an ENSO event have reached ages 2-6, those years are probably cold, and once those same pups that were born in an ENSO event have reached 7+ survival is maybe lower again because they're back into an ENSO event. 

##### Management metrics: *K*, and MNPL 

The two-step approach (e.g., Laake et al update) produces the following estimates of *K* and MNLP: approximately 292k individuals for a carrying capacity, and a MNPL of 175,364. 

```{r post hoc metrics}

print(round(summary_table))

plot(x = 1990:2024, y = recon$Ntot, type = "l", col = "darkblue", lwd = 2, 
     main = "Reconstructed population trajectory",
     ylab = "Total abundance", xlab = "")

plot(x = 1990:2024, y = implied_fecundity, type = "b", col = "darkred", pch = 16,
     # main = "Implied birth rate",
     ylab = "Fecundity (pups per adult female)", xlab = "")

```

Plots show reconstructed population trajectory (were pup counts increasing that rapidly 2000-2010?) and implied fecundity rate (is this about right for CSL? seems a bit low for the pinnipeds I'm used to, or at least more volatile). 

The integrated population model produces the following estimates: 

```{r IPM metrics}



fec_plot

DD_plot

```


##### Abundance and posterior predictive checks  


#### Discussion

Laake et al. 2018 metrics: 

- ~275,000 (K)
- MNPL: 183,481... though just based on pups, right...? 
- abundance in 2014: 258k animals, 40% above MNPL so within OSP range
- N/K DD: -1.68 (-3.78,-0.03)

#### Figures and tables 

Figure 1. Map?

```{r phi plot}

phi_plot

```
Figure 2. Posterior median and 95% credible intervals for average age-specific CSL survival probabilities over the study period. 

```{r fec plot}

fec_plot

```
Figure 3. Posterior median and 95% credible intervals for average fecundity (pooled for ages 4:9 and 10:22) probabilities over the study period.


```{r p plot}

```
Figure 4. Posterior median and 95% credible intervals for average age-specific CSL detection probabilities over the study period.

```{r N and pp plot}

```
Figure 5. Posterior median and 95% credible intervals for annual CSL abundance and predicted number of pups (color) and imputed census counts (color) over the study period.

```{r sst predictions}

sst_plot

```
Figure 6. Posterior median and 95% credible intervals of (a) age- and sex-specific logit-scale effects of sea surface temperature on CSL survival probabilities, and (b) predicted age- and sex-specific survival probabilities across the range of sea surface temperature anomalies over the study period. 


```{r MNPL curve}

```
Figure 7. Maximum net productivity level curve, showing....?

```{r table 1}

```
Table 1: Annual abundances for M and F and tot and pup multiplier over time.

```{r table 2}

```
Table 2: Median and 95% credible intervals for population model parameters, including of B_sst_a, beta.DD, K, MNPL median and 95% CI, MNPL/K, N in start/stop yrs, annual growth/max growth, and N_2024/MNPL (OSP status). 

#### Appendix

Describe model diagnostics and fit. 

```{r ppc and posterior prior overlaps}



```

