---
title: ""
output: html_document
date: ""
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = F, warning = F, eval = F, message = F, cache = F, cache.lazy = T, fig.width = 6, fig.height = 5, fig.align = 'center')

library(here) #v1.0.1
library(dplyr) #v1.0.5
# library(nimble) #v0.12.1
# library(coda)
library(ggplot2)
library(tidyverse)
library(readr)

source(here::here('scripts', 'PlotTheme.R'))

```

```{r vital rates, eval = T}

library(coda)
out <- readRDS(here::here('results', 'out_sst_probs.RDS'))

all_pars <- colnames(out[[1]])
outmat <- as.matrix(out)

post_sum <- data.frame(
  med = apply(outmat, 2, function(x) quantile(x, probs = 0.5, na.rm = T, names = F)),
  lower = apply(outmat, 2, function(x) quantile(x, probs = 0.025, na.rm = T, names = F)),
  upper = apply(outmat, 2, function(x) quantile(x, probs = 0.975, na.rm = T, names = F)))
post_sum$variable <- row.names(post_sum)

#mean rates
# sex.pars <- all_pars[which(grepl('b.sex', all_pars))]
mu.pars <- all_pars[which(grepl('mu.', all_pars))]

# b.mass_sum <- post_sum %>% filter(grepl('b.mass', variable)) %>%
#   # transform(var_name = ifelse(grepl('b.mass.f', variable), '1',
#   #                                    ifelse(grepl('b.mass.m', variable), '1M',
#                   transform(var_name = ifelse(grepl('b.mass1', variable), '1',
#                        ifelse(grepl('b.massP', variable), 'P', NA))) %>%
#   transform(sex = ifelse(grepl('M', variable), 'M', 'F')) %>%
#   transform(var_name = factor(var_name, levels = c('P', '1')))
#   
# ggplot(b.mass_sum,
#        aes(x = var_name, y = med, col = sex, group = sex)) +
#   geom_errorbar(aes(x = var_name, ymin=lower, ymax=upper),
#                 position = position_dodge(0.5),
#                 width = 0.25, show.legend = F) +
#   geom_point(size = 0.8, position = position_dodge(0.5)) +
#   geom_hline(aes(yintercept = 0), linetype = 'dashed') +
#   xlab('Age class') + ylab(expression(paste('Logit-scale effect of handling mass'))) +
#   ggtitle('') +
#   plot_theme(legend.position = 'top', panel.border = element_rect(fill = NA),
#              plot.title = element_text(hjust = 0.5)) +
#   scale_color_manual(values = rainbow2[c(2,5,6,7,8)], name = 'Sex')


#survival
int_dat <- post_sum %>% filter(grepl('int.phi', variable)) %>%
  # bind_rows(post_sum %>% filter(grepl('int.psi', variable))) %>%
  transform(variable = gsub('int.', '', variable)) %>%
  transform(sex = ifelse(grepl('M', variable), 'M', 'F')) %>%
  transform(age = gsub('phi', '', variable)) %>%
  transform(age = gsub('M', '', age)) %>%
   transform(age = factor(age, levels = c('P', 'PM', '1', 
                                                   '1M', '2.3', '2.3M',
                            '4.6', '4.6M', '7.9', '7.9M', '10.12', '10.12M',
                            '13.16', '13.16M', 'A', 'AM'))) 
  
  
phi_plot <- ggplot(int_dat %>% filter(grepl('phi', variable)),
       aes(age, med, color = sex)) +
  geom_errorbar(aes(x = age, ymin=lower, ymax=upper),
                position = position_dodge(0.5),
                width = 0.35) +
  geom_point(size = 0.8, position = position_dodge(0.5)) +
  xlab('Age class') + ylab(expression(paste('Survival probability ',  '(', phi, ')'))) +
  ggtitle('') +
  # facet_wrap(~sex, scales = 'free_x', drop = T) +
  plot_theme(legend.position = 'top', 
             panel.border = element_rect(fill = NA),
             plot.title = element_text(hjust = 0.5)) +
  scale_color_manual(values = c('black', 'darkgrey'), name = 'Sex')
  # scale_color_manual(values = rainbow2[c(2,6,7,8)], name = 'Sex')

#fecundity
fec_dat <- post_sum %>% filter(grepl('int.fec', variable)) %>%
  # bind_rows(post_sum %>% filter(grepl('int.psi', variable))) %>%
  transform(variable = ifelse(grepl('low', variable), 'Fec_low', 'Fec_high')) %>%
   transform(age = factor(variable, levels = c('Fec_low', 'Fec_high'),
                          labels = c('4:9', '10:22'))) 

fec_plot <- ggplot(fec_dat, aes(x = age, y = value)) +
  geom_errorbar(aes(x = age, ymin=lower, ymax=upper),
                position = position_dodge(0.5),
                width = 0.35) +
  geom_point(size = 0.8, position = position_dodge(0.5)) +
  xlab('Fecundity age bins') + ylab(expression(paste('Fecundity ',  '(', f, ')'))) +
  ggtitle('') +
  # facet_wrap(~sex, scales = 'free_x', drop = T) +
  plot_theme(legend.position = 'top', 
             panel.border = element_rect(fill = NA),
             plot.title = element_text(hjust = 0.5))

#detection
int_dat_p <- post_sum %>% filter(grepl('int.p', variable)) %>%
  # bind_rows(post_sum %>% filter(grepl('int.psi', variable))) %>%
  transform(variable = gsub('int.', '', variable)) %>%
  filter(!grepl('phi', variable)) %>%
    transform(sex = ifelse(grepl('M', variable), 'M', 'F')) %>%
  transform(age = gsub('p', '', variable)) %>%
  transform(age = gsub('M', '', age)) %>%
    transform(age = factor(age, levels = c('1', '2.3', '4.6',  
                                                   '7.12', 'A'))) 

p_plot <- ggplot(int_dat_p,
       aes(age, med, color = sex)) +
  geom_errorbar(aes(x = age, ymin=lower, ymax=upper),
                position = position_dodge(0.5),
                width = 0.25) +
  geom_point(size = 0.8, position = position_dodge(0.5)) +
  xlab('Age class') + ylab(expression(paste('Detection probability ',  '(', p, ')'))) +
  ggtitle('') +
  # facet_wrap(~sex, scales = 'free_x', drop = T) +
  plot_theme(legend.position = 'top', 
             panel.border = element_rect(fill = NA),
             plot.title = element_text(hjust = 0.5)) +
  scale_color_manual(values = c('black', 'darkgrey'), name = 'Sex')
  # scale_color_manual(values = rainbow2[c(2,6,7,8)], name = 'Sex')

##time-varying for Sharon
phi_post_sum <- post_sum %>% filter(grepl('.prob', variable)) %>%
  filter(grepl('phi', variable)) %>%
  transform(year = as.numeric(as.character(gsub('.+\\[([0-9]+)\\].*$', '\\1',
                                                variable)))) %>%
  transform(variable = sub("\\[.*", "", variable)) %>%
  transform(variable = gsub('.prob', '', variable)) %>%
  transform(variable = factor(variable, levels = c('phiP', 'phiPM', 'phi1', 'phi1M', 'phi2.3', 'phi2.3M',
                            'phi4.6', 'phi4.6M', 'phi7.9', 'phi7.9M', 'phi10.12', 'phi10.12M',
                            'phi13.16', 'phi13.16M', 'phiA', 'phiAM'))) %>%
  transform(var_name = gsub('M', '', variable)) %>%
  transform(var_name = gsub('phi', '', var_name)) %>%
  transform(sex = ifelse(grepl('M', variable), 'Male', 'Female')) %>%
  filter(sex != 'Male')
  
plot_dat <- phi_post_sum %>%
  transform(age = factor(var_name, levels = c('P', '1', '2.3', '4.6', '7.9', '10.12', '13.16', 'A'),
                         labels = c('0-1', '1-2', '2-4', '4-7', '7-10', '10-13', '13-17', '17+'))) #%>%
  # filter(age %in% c('0-1', '1-2', '2-4', '))


phit_plot <- ggplot(plot_dat, aes(year, med, col = age)) +
  geom_errorbar(aes(x = year, ymin=lower, ymax=upper), position = position_dodge(0.5),
                width = 0.5, show.legend = F, alpha = 0.5) +
  geom_point(size = 0.8, position = position_dodge(0.5)) +
  geom_line(linewidth = 0.7, position = position_dodge(0.5)) +
  xlab('') + ylab(expression(paste('Survival probability'))) +
  ggtitle('') +
  # facet_wrap(~var_name, nrow = 2) +
  plot_theme(legend.position = 'top', panel.border = element_rect(fill = NA),
             plot.title = element_text(hjust = 0.5)) +
  # guides(color = guide_legend("", nrow = 1, byrow = T)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  scale_color_manual(values = rainbow2, name = 'Age') +
  scale_x_continuous(breaks = c(seq(1, 35, by = 2)), labels = c(seq(1990, 2024, by = 2)))

```

```{r covariate effects and predictions, eval = T}

sst_raw <- read.csv(here('data', '2025_CCEP_sst.csv'), stringsAsFactors = F, header = T)

#from S. Melin in project summary notes docx
#use pupOcttoJune: pup survival from marking to 1st birthday
#use pupAprtoSept: pup survival via gestational foraging success/pup weight effects
#use nonpupJulytoJune: survival of non-pup age classes - sea lion "year"

sst_dat <- sst_raw %>% filter(time > 1989) #%>% filter(time > 2010)

#sst
b.sst_sum <- post_sum %>% filter(grepl('b.sst', variable)) %>%
  # transform(var_name = ifelse(grepl('b.mass.f', variable), '1', 
  #                                    ifelse(grepl('b.mass.m', variable), '1M',
                  transform(var_name = gsub('b.sst.', '', variable)) %>%
  transform(var_name = factor(var_name, levels = c('P', '1', '2.3', '4.6', '7.9', 'A')))

sst_eff_plot <- ggplot(b.sst_sum, aes(x = var_name, y = med)) +
  geom_errorbar(aes(x = var_name, ymin=lower, ymax=upper),
                position = position_dodge(0.5),
                width = 0.25, show.legend = F) +
  geom_point(size = 0.8, position = position_dodge(0.5)) +
  geom_hline(aes(yintercept = 0), linetype = 'dashed') +
  xlab('Age class') + ylab(expression(paste('Logit-scale effect of SST'))) +
  ggtitle('') +
  pres_theme(legend.position = 'top', #panel.border = element_rect(fill = NA),
             plot.title = element_text(hjust = 0.5))

#changes in survival from sst
## sequence of values
sst_seq <- seq(min(sst_dat[,2]), max(sst_dat[,2]), length.out = 100)
sst_np_seq <- seq(min(sst_dat[,4]), max(sst_dat[,4]), length.out = 100)

# back-transform a single MCMC row across the sequence
get_pred <- function(phi, b_sst, sequence) {
  mu <- plogis(phi)
  logit_p <- mu + b_sst*sequence
  return(1 / (1 + exp(-logit_p)))
}

samples <- as.matrix(out)
#pups
preds_p <- apply(samples, 1, function(s) {
  get_pred(s["int.phiP"], s["b.sst.P"], sst_seq)
})

preds_pm <- apply(samples, 1, function(s) {
  get_pred(s["int.phiPM"], s["b.sst.P"], sst_seq)
})

preds_1 <- apply(samples, 1, function(s) {
  get_pred(s["int.phi1"], s["b.sst.1"], sst_seq)
})

preds_1m <- apply(samples, 1, function(s) {
  get_pred(s["int.phi1M"], s["b.sst.1"], sst_seq)
})

preds_2.3 <- apply(samples, 1, function(s) {
  get_pred(s["int.phi2.3"], s["b.sst.2.3"], sst_seq)
})

preds_2.3m <- apply(samples, 1, function(s) {
  get_pred(s["int.phi2.3M"], s["b.sst.2.3M"], sst_seq)
})

preds_4.6 <- apply(samples, 1, function(s) {
  get_pred(s["int.phi4.6"], s["b.sst.4.6"], sst_seq)
})

# preds_4.6m <- apply(samples, 1, function(s) {
#   get_pred(s["int.phi4.6M"], s["b.sst.4.6M"], sst_seq)
# })

preds_7.9 <- apply(samples, 1, function(s) {
  get_pred(s["int.phi7.9"], s["b.sst.7.9"], sst_seq)
})

# preds_7.9m <- apply(samples, 1, function(s) {
#   get_pred(s["int.phi7.9M"], s["b.sst.7.9M"], sst_seq)
# })

preds_10.12 <- apply(samples, 1, function(s) {
  get_pred(s["int.phi10.12"], s["b.sst.A"], sst_seq)
})

preds_13.16 <- apply(samples, 1, function(s) {
  get_pred(s["int.phi13.16"], s["b.sst.A"], sst_seq)
})

preds_A <- apply(samples, 1, function(s) {
  get_pred(s["int.phiA"], s["b.sst.A"], sst_seq)
})

summarize_preds_pup <- function(pred_matrix) {
  data.frame(
    sst_val = sst_seq,
    val = 1:length(sst_seq),
    mean_phi = rowMeans(pred_matrix),
    low_phi  = apply(pred_matrix, 1, quantile, probs = 0.025),
    high_phi = apply(pred_matrix, 1, quantile, probs = 0.975))
}

summarize_preds_np <- function(pred_matrix) {
  data.frame(
    sst_val = sst_np_seq,
    val = 1:length(sst_np_seq),
    mean_phi = rowMeans(pred_matrix),
    low_phi  = apply(pred_matrix, 1, quantile, probs = 0.025),
    high_phi = apply(pred_matrix, 1, quantile, probs = 0.975))
}

preds_all <- rbind(summarize_preds_pup(preds_p) %>%
                     transform(age = 'P', sex = 'Fem'),
                   summarize_preds_pup(preds_pm) %>%
                     transform(age = 'P', sex = 'Male')) %>%
  bind_rows(rbind(summarize_preds_np(preds_1) %>%
                     transform(age = '1', sex = 'Fem'),
                   summarize_preds_np(preds_1m) %>%
                     transform(age = '1', sex = 'Male'))) %>%
  # bind_rows(rbind(summarize_preds_np(preds_2.3) %>%
  #                    transform(age = '2:3', sex = 'Fem'),
  #                  summarize_preds_np(preds_2.3m) %>%
  #                    transform(age = '2:3', sex = 'Male'))) %>%
  # bind_rows(rbind(summarize_preds_np(preds_4.6) %>%
  #                    transform(age = '4:6', sex = 'Fem'),
  #                  summarize_preds_np(preds_4.6m) %>%
  #                    transform(age = '4:6', sex = 'Male'))) %>%
  bind_rows(summarize_preds_np(preds_7.9) %>%
                     transform(age = '7:9', sex = 'Fem')) %>%
  bind_rows(summarize_preds_np(preds_10.12) %>%
                     transform(age = '10:12', sex = 'Fem')) %>%
  bind_rows(summarize_preds_np(preds_13.16) %>%
                     transform(age = '13:16', sex = 'Fem')) %>%
  bind_rows(summarize_preds_np(preds_A) %>%
                     transform(age = '17+', sex = 'Fem')) %>%
  transform(age = factor(age, levels = c('P', '1', '2:3', '4:6', '7:9',
                                         '10:12', '13:16', '17+')))

#includes both sexes
# sst_preds_plot <- ggplot(preds_all, aes(x = sst_val, y = mean_phi, col = sex)) +
#   geom_ribbon(aes(ymin = low_phi, ymax = high_phi), alpha = 0.2, color = NA) +
#   geom_line(linewidth = 1) +
#   facet_grid(.~age) +
#   labs(x = "SST anomalies",
#        y = "Predicted survival probability") +
#   theme_bw() +
#   theme(legend.position = 'top',
#                 axis.title.x = element_text(margin = margin(t = 7))) +
#   scale_y_continuous(limits = c(0, 1)) + 
#   scale_color_manual(values = rainbow2[c(2,6)], name = 'Sex') + 
#   scale_fill_manual(values = rainbow2[c(2,6)]) #+ 
#   # scale_x_continuous(limits = c(-1.5, 2.5), breaks = c(seq(-1.5,2.5, length = 5)),
#   #                    labels = round(c(seq(0.3,0.65, length = 5)),2))

sst_preds_plot <- ggplot(preds_all %>% filter(sex != 'M'), aes(x = sst_val, y = mean_phi, col = age)) +
  geom_ribbon(data = preds_all, aes(ymin = low_phi, ymax = high_phi, fill = age), 
              alpha = 0.2, color = NA, show.legend = F) +
  geom_line(linewidth = 1) +
  # facet_grid(.~age) +
  labs(x = "SST anomalies",
       y = "Predicted survival probability") +
  theme_bw() +
  theme(legend.position = 'top',
                axis.title.x = element_text(margin = margin(t = 7))) +
  scale_y_continuous(limits = c(0, 1)) + 
  scale_color_manual(values = rainbow2[c(2,4,6,7,8,9)], name = 'Age') + 
  scale_fill_manual(values = rainbow2[c(2,4,6,7,8,9)]) #+ 
  # scale_x_continuous(limits = c(-1.5, 2.5), breaks = c(seq(-1.5,2.5, length = 5)),
  #                    labels = round(c(seq(0.3,0.65, length = 5)),2))

#ENSO "memory" effects
out_ENSO <- readRDS(file = here('results', 'out_ENSO_memory.RDS'))

# all_pars <- colnames(out_ENSO[[1]])
outmat_ENSO <- as.matrix(out_ENSO)

post_sum_ENSO <- data.frame(
  med = apply(outmat_ENSO, 2, function(x) quantile(x, probs = 0.5, na.rm = T, names = F)),
  lower = apply(outmat_ENSO, 2, function(x) quantile(x, probs = 0.025, na.rm = T, names = F)),
  upper = apply(outmat_ENSO, 2, function(x) quantile(x, probs = 0.975, na.rm = T, names = F)))
post_sum_ENSO$variable <- row.names(post_sum_ENSO)


b.enso <- post_sum_ENSO %>% 
  filter(grepl('b.enso', variable)) %>%
  transform(var_name = gsub('b.enso.', '', variable)) %>%
  filter(med != 0) %>%
  transform(var_name = factor(var_name, levels = c('1[1]', '2.3[1]', '4.6[1]', 'A[1]'), 
                              labels = c('1', '2.3', '4.6', '7+')))

enso_eff_plot <- ggplot(b.enso, aes(x = var_name, y = med)) +
  geom_errorbar(aes(x = var_name, ymin=lower, ymax=upper),
                position = position_dodge(0.5),
                width = 0.25, show.legend = F) +
  geom_point(size = 0.8, position = position_dodge(0.5)) +
  geom_hline(aes(yintercept = 0), linetype = 'dashed') +
  xlab('Age class') + ylab(expression(paste('Logit-scale effect of ENSO birth year'))) +
  ggtitle('') +
  pres_theme(legend.position = 'top', #panel.border = element_rect(fill = NA),
             plot.title = element_text(hjust = 0.5))

```


```{r calculate metrics from MR fit, eval = T}

library(imputeTS)

pup_cnts <- read_tsv(here('data', 'CSL_San Miguel Island_PupCounts_20260622.txt')) %>%
  filter(Year > 1989)

pup_cnt_filled <- na_interpolation(pup_cnts[,1:5]) %>%
  mutate(US_PupCnt = rowSums(across(-Year)))

#stable age logic to initialize N_f and N_m
# define survival rates
n_occasions <- 36
phi_P <- 0.65
phi_1 <- 0.8
phi_J <- 0.92 # Juveniles (Ages 2-5)
phi_A <- 0.85 # Adults (Ages 6-13)
phi_AA <- 0.75
phi_AM <- 0.8
phi_AAM <- 0.5

# storage
N_f_track <- matrix(0, nrow = 25, ncol = (n_occasions-1))
N_m_track <- matrix(0, nrow = 20, ncol = (n_occasions-1))

# year 1; assuming 50:50 sex ratio for pups
N_f_track[1, 1] <- round(pup_cnt_filled$US_PupCnt[1] / 2)
N_m_track[1, 1] <- pup_cnt_filled$US_PupCnt[1] - N_f_track[1, 1]

for (a in 1:24) {
  surv <- ifelse(a == 1, phi_P, 
          ifelse(a == 2, phi_1, 
          ifelse(a >= 3 & a <= 5, phi_J,
          ifelse(a >= 6 & a <= 16, phi_A, phi_AA))))
          
  N_f_track[a + 1, 1] <- round(N_f_track[a, 1] * surv)
}

for (a in 1:19) {
  surv <- ifelse(a == 1, phi_P, 
          ifelse(a == 2, phi_1, 
          ifelse(a >= 3 & a <= 5, phi_J,
          ifelse(a >= 6 & a <= 12, phi_AM, phi_AAM))))
          
  N_m_track[a + 1, 1] <- round(N_m_track[a, 1] * surv)
}

# dynamically project the rest of the matrix forward through time
for (t in 1:(n_occasions - 2)) {
  
  # Insert the pup count for the next year
  N_f_track[1, t+1] <- round(pup_cnt_filled$US_PupCnt[t+1]/2)
  N_m_track[1, t+1] <- pup_cnt_filled$US_PupCnt[t+1] - N_f_track[1, t+1]
  
  # Pups transition to yearlings (Index 1 -> 2)
  N_f_track[2, t+1] <- round(N_f_track[1, t] * phi_P)
  N_m_track[2, t+1] <- round(N_m_track[1, t] * phi_P)
  
  # Yearlings transition to Age 2 (Index 2 -> 3)
  N_f_track[3, t+1] <- round(N_f_track[2, t] * phi_1)
  N_m_track[3, t+1] <- round(N_m_track[2, t] * phi_1)
  
  # Juveniles transition (Indices 3-5 -> 4-6)
  for (a in 3:5) {
    N_f_track[a+1, t+1] <- round(N_f_track[a, t] * phi_J)
    N_m_track[a+1, t+1] <- round(N_m_track[a, t] * phi_J)
  }
  
  # Adults transition (Indices 6-24 -> 7-25)
  for (a in 6:12) {
    N_f_track[a+1, t+1] <- round(N_f_track[a, t] * phi_A)
    N_m_track[a+1, t+1] <- round(N_m_track[a, t] * phi_AM)
  }
  
    for (a in 13:24) {
    N_f_track[a+1, t+1] <- round(N_f_track[a, t] * phi_AA)
    }
  
  for (a in 13:19) {
    N_m_track[a+1, t+1] <- round(N_m_track[a, t] * phi_AAM)
  }
}

# row 1 (pups) is assigned deterministically in the model, NA for inits
N_f_init <- N_f_track
N_m_init <- N_m_track

N_f_init[1, ] <- NA
N_m_init[1, ] <- NA

# use year 1 for for dpois() priors
init_pop_f <- N_f_track[, 1]
init_pop_m <- N_m_track[, 1]


##calculate MNPL and OSP

#simplified approach assuming 60%, or use vital rates from IPM
# MNPL_estimates <- K_estimates * 0.60

#use vital rates from IPM
# extract MCMC matrix 
outmat <- as.matrix(out)

pup_counts <- pup_cnt_filled$US_PupCnt

# Reconstruct abundance deterministically from pup counts and survival
# pup_counts Vector of observed annual pup census counts
# phi_f Matrix of female survival (rows = age classes 1:24, cols = years)
# phi_m Matrix of male survival (rows = age classes 1:24, cols = years)
# init_f Initial female age vector at t=1 (length = max_age)
# init_m Initial male age vector at t=1 (length = max_age)
init_f <- N_f_track[,1]
init_m <- N_m_track[,1]

reconstruct_pop <- function(pup_counts, phi_f, phi_m, init_f, init_m) {
  n_occasions <- length(pup_counts)
  max_age <- nrow(phi_f)
  max_age_m <- nrow(phi_m)
  
  N_f <- matrix(0, nrow = max_age, ncol = n_occasions)
  N_m <- matrix(0, nrow = max_age_m, ncol = n_occasions)
  
  # Seed Year 1 with starting age distribution
  N_f[,1] <- N_f_track[,1]
  N_m[,1] <- N_m_track[,1]
  
  for (t in 1:(n_occasions - 1)) {
    # pups (Index 1) enter directly from empirical census counts
    # Assumes a 50:50 sex ratio at birth
    N_f[1,t+1] <- pup_counts[t+1]/2
    N_m[1,t+1] <- pup_counts[t+1]/2
    
    # age up older cohorts deterministically using CJS survival rates
    for (a in 1:(max_age - 1)) {
      N_f[a+1,t+1] <- N_f[a,t]*phi_f[a,t]
    }
  
  for (a in 1:(max_age_m - 1)) {
      N_m[a+1,t+1] <- N_m[a,t]*phi_m[a,t]
    }
  }
  
  # Calculate total abundance per year 
  Ntot <- colSums(N_f) + colSums(N_m)
  
  return(list(Ntot = Ntot, N_f = N_f, N_m = N_m, pups = pup_counts))
}

# Fit Theta-Logistic Model to reconstructed abundance via Sum of Squared Errors
# N_series is a vector of reconstructed total population size over time
fit_theta_logistic <- function(N_series) {
  n_occasions <- length(N_series)
  
  # Objective function: Sum of Squared Errors (SSE)
  sse_fn <- function(params) {
    r     <- params[1]
    K     <- params[2]
    theta <- params[3]
    
    N_pred <- numeric(n_occasions)
    N_pred[1] <- N_series[1]
    
    for (t in 1:(n_occasions - 1)) {
      # Discrete theta-logistic step
      growth <- r * N_pred[t] * (1 - (N_pred[t] / K)^theta)
      N_pred[t+1] <- N_pred[t] + growth
      
      # Penalty for numerical overflow or negative population dips
      if (is.na(N_pred[t+1]) || N_pred[t+1] <= 0) return(1e10)
    }
    
    return(sum((N_series - N_pred)^2))
  }
  
  # Set reasonable biological starting bounds for optimization
  init_K     <- max(N_series) * 1.1
  init_r     <- 0.07 # Standard pinniped intrinsic growth rate
  init_theta <- 2.0  # Standard delayed density dependence
  
  opt <- optim(
    par = c(r = init_r, K = init_K, theta = init_theta),
    fn = sse_fn,
    method = "L-BFGS-B",
    lower = c(r = 0.01, K = max(N_series) * 0.8, theta = 0.5),
    upper = c(r = 0.20, K = max(N_series) * 3.0, theta = 10.0)
  )
  
  # Extract estimates and analytically calculate MNPL
  r_est     <- opt$par["r"]
  K_est     <- opt$par["K"]
  # theta_est <- opt$par["theta"]
  theta_est <- 2.4 #MMPA standard; not always possible to estimate r and K and theta
  mnpl_est  <- K_est * ((1 / (theta_est + 1))^(1 / theta_est))
  
  return(data.frame(
    r = r_est, 
    K = K_est, 
    theta = theta_est, 
    MNPL = mnpl_est, 
    MNPL_ratio = mnpl_est / K_est,
    Convergence = opt$convergence
  ))
}

# loop across CJS posterior draws)

out <- readRDS(here::here('results', 'out_sst_probs.RDS'))
outmat <- as.matrix(out)

nsim <- 500
results_list <- vector("list", nsim)
n_occasions <- 35

phiPcols <- paste0('phiP.prob[', 1:n_occasions, ']')
phi1cols <- paste0('phi1.prob[', 1:n_occasions, ']')
phi2.3cols <- paste0('phi2.3.prob[', 1:n_occasions, ']')
phi4.6cols <- paste0('phi4.6.prob[', 1:n_occasions, ']')
phi7.9cols <- paste0('phi7.9.prob[', 1:n_occasions, ']')
phi10.12cols <- paste0('phi10.12.prob[', 1:n_occasions, ']')
phi13.16cols <- paste0('phi13.16.prob[', 1:n_occasions, ']')
phiAcols <- paste0('phiA.prob[', 1:n_occasions, ']')

phiPMcols <- paste0('phiPM.prob[', 1:n_occasions, ']')
phi1Mcols <- paste0('phi1M.prob[', 1:n_occasions, ']')
phi2.3Mcols <- paste0('phi2.3M.prob[', 1:n_occasions, ']')
phi4.6Mcols <- paste0('phi4.6M.prob[', 1:n_occasions, ']')
phi7.9Mcols <- paste0('phi7.9M.prob[', 1:n_occasions, ']')
phi10.12Mcols <- paste0('phi10.12M.prob[', 1:n_occasions, ']')
phi13.16Mcols <- paste0('phi13.16M.prob[', 1:n_occasions, ']')
phiAMcols <- paste0('phiAM.prob[', 1:n_occasions, ']')

# cat("Running post-hoc reconstruction across", nsim, "CJS draws...\n")
for (k in 1:nsim) {

  phiP <- outmat[k,phiPcols]
  phi1 <- outmat[k,phi1cols]
  phi2.3 <- outmat[k,phi2.3cols]
  phi4.6 <- outmat[k,phi4.6cols]
  phi7.9 <- outmat[k,phi7.9cols]
  phi10.12 <- outmat[k,phi10.12cols]
  phi13.16 <- outmat[k,phi13.16cols]
  phiA <- outmat[k,phiAcols]
  
  phiPM <- outmat[k,phiPMcols]
  phi1M <- outmat[k,phi1Mcols]
  phi2.3M <- outmat[k,phi2.3Mcols]
  phi4.6M <- outmat[k,phi4.6Mcols]
  phi7.9M <- outmat[k,phi7.9Mcols]
  phi10.12M <- outmat[k,phi10.12Mcols]
  phi13.16M <- outmat[k,phi13.16Mcols]
  phiAM <- outmat[k,phiAMcols]
  
  phi_f <- rbind(phiP, phi1, phi2.3, phi2.3, phi4.6, phi4.6, phi4.6, 
                 phi7.9, phi7.9, phi7.9, phi10.12, phi10.12, phi10.12, 
                 phi13.16, phi13.16, phi13.16, phi13.16,
                 phiA, phiA, phiA, phiA, phiA, phiA, phiA, phiA)
  phi_m <- rbind(phiPM, phi1M, phi2.3M, phi2.3M, phi4.6M, phi4.6M, phi4.6M,
                 phi7.9M, phi7.9M, phi7.9M, phi10.12M, phi10.12M, phi10.12M, 
                 phi13.16M, phi13.16M, phi13.16M, phi13.16M, 
                 phiAM, phiAM, phiAM)

  # Run deterministic projection
  recon <- reconstruct_pop(pup_counts, phi_f, phi_m, init_f, init_m)
  
  # Fit theta-logistic model
  fit <- fit_theta_logistic(recon$Ntot)
  fit$draw <- k
  results_list[[k]] <- fit
}

# Combine results into a final summary dataframe
post <- bind_rows(results_list)

# View summary statistics 
summary_table <- post %>%
  summarise(
    K_median    = median(K),
    K_LCI       = quantile(K, 0.025),
    K_UCI       = quantile(K, 0.975),
    MNPL_median = median(MNPL),
    MNPL_LCI    = quantile(MNPL, 0.025),
    MNPL_UCI    = quantile(MNPL, 0.975),
    Theta_median = median(theta),
    Theta_LCI = quantile(theta, 0.025),
    theta_UCI = quantile(theta, 0.975)
  )

# print(summary_table)


# calculate the implied fecundity
# How many pups were born per adult female (assuming ages 4-24 are adults)?
# adult_females <- colSums(recon$N_f[4:24, ])
# implied_fecundity_low <- pup_counts / adult_females

fem_5.7 <- colSums(recon$N_f[6:8, ])
fem_8.17 <- colSums(recon$N_f[9:18, ])
fem_18 <- colSums(recon$N_f[19:24, ])

implied_fec_5.7 <- pup_counts/fem_5.7
implied_fec_8.17 <- pup_counts/fem_8.17
implied_fec_18 <- pup_counts/fem_18

fec_long <- data.frame(fec5.7 = implied_fec_5.7,
                       fec8.17 = implied_fec_8.17,
                       fec18 = implied_fec_18,
                       yr = 1990:2024)

# Ntot_recon_plot <- plot(x = 1990:2024, y = recon$Ntot, type = "l", col = "darkblue", lwd = 2, 
#      main = "Reconstructed population trajectory",
#      ylab = "Total abundance", xlab = "")
# 
# fec_calc_plot <- plot(x = 1990:2024, y = implied_fecundity, type = "b", col = "darkred", pch = 16,
#      # main = "Implied birth rate",
#      ylab = "Fecundity (pups per adult female)", xlab = "")


```

```{r calculate metrics from IPM fit and post pred checks, eval = T, results = 'keep'}

out_IPM <- readRDS(file = here('results', 'out_IPM_sst_DD_slice2.RDS'))
outmat <- do.call(rbind, out_IPM)

outmat <- as.matrix(out_IPM)

post_sumIPM <- data.frame(
  med = apply(outmat, 2, function(x) quantile(x, probs = 0.5, na.rm = T, names = F)),
  lower = apply(outmat, 2, function(x) quantile(x, probs = 0.025, na.rm = T, names = F)),
  upper = apply(outmat, 2, function(x) quantile(x, probs = 0.975, na.rm = T, names = F)))
post_sumIPM$variable <- row.names(post_sumIPM)

#survival
int_datIPM <- post_sumIPM %>% filter(grepl('int.phi', variable)) %>%
  # bind_rows(post_sum %>% filter(grepl('int.psi', variable))) %>%
  transform(variable = gsub('int.', '', variable)) %>%
  transform(sex = ifelse(grepl('M', variable), 'M', 'F')) %>%
  transform(age = gsub('phi', '', variable)) %>%
  transform(age = gsub('M', '', age)) %>%
   transform(age = factor(age, levels = c('P', 'PM', '1', 
                                                   '1M', '2.3', '2.3M',
                            '4.6', '4.6M', '7.9', '7.9M', '10.12', '10.12M',
                            '13.16', '13.16M', 'A', 'AM'))) %>%
  transform(mod = 'IPM') %>%
  bind_rows(int_dat %>% transform(mod = 'MR'))
  
plot_dat <- int_datIPM %>% filter(sex != 'M')

phi_plot_compare <- ggplot(plot_dat, aes(age, med, color = mod)) +
  geom_errorbar(aes(x = age, ymin=lower, ymax=upper),
                position = position_dodge(0.5),
                width = 0.35) +
  geom_point(size = 0.8, position = position_dodge(0.5)) +
  xlab('Age class') + ylab(expression(paste('Survival probability ',  '(', phi, ')'))) +
  ggtitle('') +
  # facet_wrap(~sex, scales = 'free_x', drop = T) +
  plot_theme(legend.position = 'top', 
             panel.border = element_rect(fill = NA),
             plot.title = element_text(hjust = 0.5)) +
  scale_color_manual(values = c('black', 'darkgrey'), name = 'Model')

#fecundity
fec_dat <- data.frame(outmat) %>% dplyr::select(contains('int.fec'))
colnames(fec_dat) <- c('Age 10+', 'Age 4-9')

fec_plot <- ggplot(fec_dat %>% reshape2::melt() %>%
                     transform(variable = factor(variable, levels = c('Age 4-9', 'Age 10+'))), 
                   aes(x = variable, y = value)) +
  geom_violin() +
  ylab('Fecundity') + xlab('') +
  theme_bw()

DD_sum <- post_sumIPM %>% filter(grepl('beta.DD', variable))

DD_plot <- ggplot(DD_sum, aes(y = med, x = variable)) +
  geom_point(aes(x = variable, y = med)) +
  geom_errorbar(aes(x = variable, ymin = lower, ymax = upper), width = 0.1) +
  ylab('Density dependence') + xlab('') +
  theme_bw()

##PPC
# expected pups columns chronologically
pup_cols <- paste0("expected_pups_tot[", 2:n_occasions, "]") # Starts at 2 because t+1 in the loop
pup_post <- outmat[, pup_cols]

# calculate the median and 95% CI for each year and append census data 
pup_summary <- data.frame(
  TimeStep = 2:n_occasions, 
  year = 1991:2024,
  Expected_Median = apply(pup_post, 2, median),
  Expected_LCI = apply(pup_post, 2, quantile, probs = 0.025),
  Expected_UCI = apply(pup_post, 2, quantile, probs = 0.975),
  Observed = pup_cnt_filled$US_PupCnt[2:n_occasions]
)

pups_ppc_plot <- ggplot(pup_summary, aes(x = year)) +
  geom_ribbon(aes(ymin = Expected_LCI, ymax = Expected_UCI), 
              fill = "steelblue", alpha = 0.3) +
  geom_line(aes(y = Expected_Median), color = "steelblue", size = 1) +
  geom_point(aes(y = Observed), color = "black", size = 2) +
  geom_line(aes(y = Observed), linetype = 'dashed', col = 'darkgrey') +
  theme_bw() +
  labs(
    title = "Posterior predictive check",
    # subtitle = "Black dots = Census Data | Blue line & ribbon = Model Expected Pups (95% CI)",
    x = "",
    y = "Total pups") +
  plot_theme()


#### also plot scatterplot of int.phiP and beta.DD?




##calculate K and MNPL: theoretical K
# Number of posterior draws to simulate (e.g., sample 1000 to save time)
n_draws <- 500#0
draw_indices <- sample(1:nrow(outmat), n_draws)

# storage
theo_K_estimates <- numeric(n_draws)
N_tot_tracker <- numeric()

# Run the simulation for each sampled posterior draw
for (i in 1:n_draws) {
  row_idx <- draw_indices[i]

  # extract vital rates for each MCMC draw
  b_sst_P <- outmat[row_idx, 'b.sst.P']
  int_phiP   <- outmat[row_idx, "int.phiP"]
  beta_DD <- outmat[row_idx, "beta.DD"]
  int_phi1   <- outmat[row_idx, "int.phi1"]
  int_phi2.3   <- outmat[row_idx, "int.phi2.3"]
  int_phi4.6   <- outmat[row_idx, "int.phi4.6"]
  int_phi7.9   <- outmat[row_idx, "int.phi7.9"]
  int_phi10.12   <- outmat[row_idx, "int.phi10.12"]
  int_phi13.16   <- outmat[row_idx, "int.phi13.16"]
  int_phiA   <- outmat[row_idx, "int.phiA"]
  fec_low  <- outmat[row_idx, "int.fec_low"]
  fec_high <- outmat[row_idx, "int.fec_high"]

  int_phiPM   <- outmat[row_idx, "int.phiPM"]
  int_phi1M     <- outmat[row_idx, "int.phi1M"]
  int_phi2.3M   <- outmat[row_idx, "int.phi2.3M"]
  int_phi4.6M   <- outmat[row_idx, "int.phi4.6M"]
  int_phi7.9M   <- outmat[row_idx, "int.phi7.9M"]
  int_phi10.12M <- outmat[row_idx, "int.phi10.12M"]
  int_phi13.16M <- outmat[row_idx, "int.phi13.16M"]
  int_phiAM     <- outmat[row_idx, "int.phiAM"]

  # convert baseline survival to probability scale (assuming SST = 0)
  mu_P <- qlogis(int_phiP)
  mu_1 <- qlogis(int_phi1)
  mu_2.3 <- qlogis(int_phi2.3)
  mu_4.6 <- qlogis(int_phi4.6)
  mu_7.9 <- qlogis(int_phi7.9)
  mu_10.12 <- qlogis(int_phi10.12)
  mu_13.16 <- qlogis(int_phi13.16)
  mu_A <- qlogis(int_phiA)
  mu_PM <- qlogis(int_phiPM)
  mu_1M <- qlogis(int_phi1M)
  mu_2.3M <- qlogis(int_phi2.3M)
  mu_4.6M <- qlogis(int_phi4.6M)
  mu_7.9M <- qlogis(int_phi7.9M)
  mu_10.12M <- qlogis(int_phi10.12M)
  mu_13.16M <- qlogis(int_phi13.16M)
  mu_AM <- qlogis(int_phiAM)

  # initialize a starting population vector
  # (e.g., 50 females per age class to the simulation)
  Nf_vec <- rep(50, 25)
  Nm_vec <- rep(50, 20)

  # project forward to reach equilibrium (K)
  for (y in 1:500) { # 200 years long enough?

    N_tot <- sum(Nf_vec) + sum(Nm_vec)

    N_tot_tracker[y] <- N_tot

    # calculate density-dependent pup survival for THIS year
    # scale N as in the model
    #average SST always
    phiP_current <- plogis(mu_P-beta_DD*((N_tot)/100000))
    phiPM_current <- plogis(mu_PM-beta_DD*((N_tot)/100000))

    # #heatwaves
    # phiP_current <- plogis(mu_P + (b_sst_P*1.5) - beta_DD*((N_tot)/100000))
    # phiPM_current <- plogis(mu_PM + (b_sst_P*1.5) - beta_DD*((N_tot)/100000))

    # calculate total pups born
    pups_born <- (sum(Nf_vec[5:10])*fec_low) + (sum(Nf_vec[11:23])*fec_high)

    # Create the next year's population vector
    Nm_vec_next <- numeric(20)
    Nm_vec_next[1] <- pups_born*0.5
    Nm_vec_next[2] <- Nm_vec[1] * phiPM_current
    Nm_vec_next[3] <- Nm_vec[2] * int_phi1M
    Nm_vec_next[4] <- Nm_vec[3] * int_phi2.3M
    Nm_vec_next[5] <- Nm_vec[4] * int_phi2.3M
    Nm_vec_next[6] <- Nm_vec[5] * int_phi4.6M
    Nm_vec_next[7] <- Nm_vec[6] * int_phi4.6M
    Nm_vec_next[8] <- Nm_vec[7] * int_phi4.6M
    Nm_vec_next[9] <- Nm_vec[8] * int_phi7.9M
    Nm_vec_next[10] <- Nm_vec[9] * int_phi7.9M
    Nm_vec_next[11] <- Nm_vec[10] * int_phi7.9M
    Nm_vec_next[12] <- Nm_vec[11] * int_phi10.12M
    Nm_vec_next[13] <- Nm_vec[12] * int_phi10.12M
    Nm_vec_next[14] <- Nm_vec[13] * int_phi10.12M
    Nm_vec_next[15] <- Nm_vec[14] * int_phi13.16M
    Nm_vec_next[16] <- Nm_vec[15] * int_phi13.16M
    Nm_vec_next[17] <- Nm_vec[16] * int_phi13.16M
    Nm_vec_next[18] <- Nm_vec[17] * int_phi13.16M

    Nf_vec_next <- numeric(25)
    Nf_vec_next[1] <- pups_born*0.5
    Nf_vec_next[2] <- Nf_vec[1] * phiP_current
    Nf_vec_next[3] <- Nf_vec[2] * int_phi1
    Nf_vec_next[4] <- Nf_vec[3] * int_phi2.3
    Nf_vec_next[5] <- Nf_vec[4] * int_phi2.3
    Nf_vec_next[6] <- Nf_vec[5] * int_phi4.6
    Nf_vec_next[7] <- Nf_vec[6] * int_phi4.6
    Nf_vec_next[8] <- Nf_vec[7] * int_phi4.6
    Nf_vec_next[9] <- Nf_vec[8] * int_phi7.9
    Nf_vec_next[10] <- Nf_vec[9] * int_phi7.9
    Nf_vec_next[11] <- Nf_vec[10] * int_phi7.9
    Nf_vec_next[12] <- Nf_vec[11] * int_phi10.12
    Nf_vec_next[13] <- Nf_vec[12] * int_phi10.12
    Nf_vec_next[14] <- Nf_vec[13] * int_phi10.12
    Nf_vec_next[15] <- Nf_vec[14] * int_phi13.16
    Nf_vec_next[16] <- Nf_vec[15] * int_phi13.16
    Nf_vec_next[17] <- Nf_vec[16] * int_phi13.16
    Nf_vec_next[18] <- Nf_vec[17] * int_phi13.16

    # absorbing state (ages 17-23)
    for (a in 18:24) {
      Nm_vec_next[a+1] <- Nm_vec[a] * int_phiAM
    }

    for (a in 18:19) {
      Nf_vec_next[a+1] <- Nf_vec[a] * int_phiA
    }

    # update population
    Nm_vec <- Nm_vec_next
    Nf_vec <- Nf_vec_next
  }

  # if (i == 1) {
  #   plot(1:500, N_tot_tracker, type = "l", col = "purple", lwd = 2,
  #        main = "Diagnostic: Projection to K",
  #        xlab = "Projection Year", ylab = "Simulated abundance")
  # }

  # store stabilized population size
  theo_K_estimates[i] <- sum(Nm_vec) + sum(Nf_vec)
}

# mean
theo_K <- mean(theo_K_estimates)
theo_K_CI <- quantile(theo_K_estimates, probs = c(0.025, 0.975), names = F)

# print(paste("Estimated theoretical K:", round(mean_K), "95% CI:", round(K_CI[1]), "-", round(K_CI[2])))

  
# ##K across a temperature gradient
# sst_scenarios <- seq(0.0, 2.0, by = 0.2)
# mean_K_by_sst <- numeric(length(sst_scenarios))
# K_ci_lower    <- numeric(length(sst_scenarios))
# K_ci_upper    <- numeric(length(sst_scenarios))
# 
# n_draws <- 100 
# draw_indices <- sample(1:nrow(outmat), n_draws)
# 
# for (s in seq_along(sst_scenarios)) {
#   target_sst <- sst_scenarios[s]
#   K_estimates_sst <- numeric(n_draws)
#   
#   for (i in 1:n_draws) {
#     row_idx <- draw_indices[i]
#     
#     # Extract vital rates and SST coefficients
#     int_phiP     <- outmat[row_idx, "int.phiP"]
#     b_sst_P      <- outmat[row_idx, "b.sst.P"] 
#     b_sst_1      <- outmat[row_idx, "b.sst.1"]
#     b_sst_2.3      <- outmat[row_idx, "b.sst.2.3"]
#     beta_DD      <- outmat[row_idx, "beta.DD"]
#     
#     int_phi1     <- outmat[row_idx, "int.phi1"]
#     int_phi2.3   <- outmat[row_idx, "int.phi2.3"]
#     int_phi4.6   <- outmat[row_idx, "int.phi4.6"]
#     int_phi7.9   <- outmat[row_idx, "int.phi7.9"]
#     int_phi10.12 <- outmat[row_idx, "int.phi10.12"]
#     int_phi13.16 <- outmat[row_idx, "int.phi13.16"]
#     int_phiA     <- outmat[row_idx, "int.phiA"]
#     fec_low      <- outmat[row_idx, "int.fec_low"]
#     fec_high     <- outmat[row_idx, "int.fec_high"]
#     
#     # Males
#     int_phiPM    <- outmat[row_idx, "int.phiPM"]
#     int_phi1M    <- outmat[row_idx, "int.phi1M"]
#     int_phi2.3M  <- outmat[row_idx, "int.phi2.3M"]
#     int_phi4.6M  <- outmat[row_idx, "int.phi4.6M"]
#     int_phi7.9M  <- outmat[row_idx, "int.phi7.9M"]
#     int_phi10.12M<- outmat[row_idx, "int.phi10.12M"]
#     int_phi13.16M<- outmat[row_idx, "int.phi13.16M"]
#     int_phiAM    <- outmat[row_idx, "int.phiAM"]
# 
#     mu_P  <- qlogis(int_phiP)
#     mu_1  <- qlogis(int_phi1)
#   mu_2.3 <- qlogis(int_phi2.3)
#   mu_4.6 <- qlogis(int_phi4.6)
#   mu_7.9 <- qlogis(int_phi7.9)
#   mu_10.12 <- qlogis(int_phi10.12)
#   mu_13.16 <- qlogis(int_phi13.16)
#   mu_A <- qlogis(int_phiA)
#    mu_PM <- qlogis(int_phiPM)
#   mu_1M <- qlogis(int_phi1M)
#   mu_2.3M <- qlogis(int_phi2.3M)
#   mu_4.6M <- qlogis(int_phi4.6M)
#   mu_7.9M <- qlogis(int_phi7.9M)
#   mu_10.12M <- qlogis(int_phi10.12M)
#   mu_13.16M <- qlogis(int_phi13.16M)
#   mu_AM <- qlogis(int_phiAM)
#     
#     Nf_vec <- rep(50, 25)
#     Nm_vec <- rep(50, 25)
# 
#     for (y in 1:400) {
#       N_tot <- sum(Nf_vec) + sum(Nm_vec)
#       
#       phiP_current  <- plogis(mu_P  + (b_sst_P * target_sst) - beta_DD*(N_tot/100000))
#       phiPM_current <- plogis(mu_PM + (b_sst_P * target_sst) - beta_DD*(N_tot/100000))
#       phi1_current  <- plogis(mu_1  + (b_sst_1 * target_sst)) 
#       phi1M_current  <- plogis(mu_1M  + (b_sst_1 * target_sst)) 
#       phi2.3_current  <- plogis(mu_2.3  + (b_sst_2.3 * target_sst)) 
#       phi2.3M_current  <- plogis(mu_2.3M  + (b_sst_2.3 * target_sst)) 
#       
#       pups_born <- (sum(Nf_vec[5:10])*fec_low) + (sum(Nf_vec[11:23])*fec_high)
# 
#       Nm_vec_next <- numeric(25)
#       Nf_vec_next <- numeric(25)
#       
#       Nm_vec_next[1] <- pups_born * 0.5
#       Nf_vec_next[1] <- pups_born * 0.5
#       
#       Nm_vec_next[2] <- Nm_vec[1] * phiPM_current
#       Nf_vec_next[2] <- Nf_vec[1] * phiP_current
#       
#       Nm_vec_next[3] <- Nm_vec[2] * phi1M_current 
#       Nf_vec_next[3] <- Nf_vec[2] * phi1_current
#       
#       Nm_vec_next[4] <- Nm_vec[3] * phi2.3M_current 
#       Nm_vec_next[5] <- Nm_vec[4] * phi2.3M_current 
#       Nf_vec_next[4] <- Nf_vec[3] * phi2.3_current
#       Nf_vec_next[5] <- Nf_vec[4] * phi2.3_current
#       
#     Nf_vec_next[6] <- Nf_vec[5] * int_phi4.6
#     Nf_vec_next[7] <- Nf_vec[6] * int_phi4.6
#     Nf_vec_next[8] <- Nf_vec[7] * int_phi4.6
#     Nf_vec_next[9] <- Nf_vec[8] * int_phi7.9
#     Nf_vec_next[10] <- Nf_vec[9] * int_phi7.9
#     Nf_vec_next[11] <- Nf_vec[10] * int_phi7.9
#     Nf_vec_next[12] <- Nf_vec[11] * int_phi10.12
#     Nf_vec_next[13] <- Nf_vec[12] * int_phi10.12
#     Nf_vec_next[14] <- Nf_vec[13] * int_phi10.12
#     Nf_vec_next[15] <- Nf_vec[14] * int_phi13.16
#     Nf_vec_next[16] <- Nf_vec[15] * int_phi13.16
#     Nf_vec_next[17] <- Nf_vec[16] * int_phi13.16
#     Nf_vec_next[18] <- Nf_vec[17] * int_phi13.16
#     
#     Nm_vec_next[6] <- Nm_vec[5] * int_phi4.6M
#     Nm_vec_next[7] <- Nm_vec[6] * int_phi4.6M
#     Nm_vec_next[8] <- Nm_vec[7] * int_phi4.6M
#     Nm_vec_next[9] <- Nm_vec[8] * int_phi7.9M
#     Nm_vec_next[10] <- Nm_vec[9] * int_phi7.9M
#     Nm_vec_next[11] <- Nm_vec[10] * int_phi7.9M
#     Nm_vec_next[12] <- Nm_vec[11] * int_phi10.12M
#     Nm_vec_next[13] <- Nm_vec[12] * int_phi10.12M
#     Nm_vec_next[14] <- Nm_vec[13] * int_phi10.12M
#     Nm_vec_next[15] <- Nm_vec[14] * int_phi13.16M
#     Nm_vec_next[16] <- Nm_vec[15] * int_phi13.16M
#     Nm_vec_next[17] <- Nm_vec[16] * int_phi13.16M
#     Nm_vec_next[18] <- Nm_vec[17] * int_phi13.16M
#       
#       for (a in 18:24) {
#         Nm_vec_next[a+1] <- Nm_vec[a] * int_phiAM
#         Nf_vec_next[a+1] <- Nf_vec[a] * int_phiA
#       }
# 
#       Nm_vec <- Nm_vec_next
#       Nf_vec <- Nf_vec_next
#     }
#     
#     K_estimates_sst[i] <- sum(Nm_vec) + sum(Nf_vec)
#   }
#   
#   mean_K_by_sst[s] <- mean(K_estimates_sst)
#   K_ci_lower[s]    <- quantile(K_estimates_sst, 0.025)
#   K_ci_upper[s]    <- quantile(K_estimates_sst, 0.975)
# }
# 
# plot(sst_scenarios, mean_K_by_sst, type = "b", pch = 16, col = "darkblue", lwd = 2,
#      ylim = c(min(K_ci_lower), max(K_ci_upper)),
#      main = "Dynamic K across SST regimes",
#      xlab = "SST anomaly (°C)", ylab = "Estimated carrying capacity (K)")
# 
# # Add confidence intervals
# lines(sst_scenarios, K_ci_lower, lty = 2, col = "blue")
# lines(sst_scenarios, K_ci_upper, lty = 2, col = "blue")
# grid()

#### more realistic sst scenarios
n_draws <- 500 
draw_indices <- sample(1:nrow(outmat), n_draws)

n_years <- 600

# Scenario a: an illustrative cyclic pattern (1.5C heatwaves for 2 yrs, every 7 yrs)
# sst_cyclic <- rep(0, n_years) # Baseline average
# for (y in 1:n_years) {
#   if (y %% 7 == 0 || y %% 7 == 1) { 
#     sst_cyclic[y] <- 1.5 
#   }
# }

# Scenario b: bootstrapped historical data
sst <- sst_dat[,3]
sst_boot <- sample(sst, n_years, replace = TRUE)

# choose scenario
# sst_vec <- sst_cyclic 
sst_vec <- sst_boot 

pop_trajectories <- matrix(0, nrow = n_draws, ncol = n_years)
realized_K <- numeric(n_draws)
mnpl_estimates <- numeric(n_draws)
mnpl_ratio <- numeric(n_draws)

for (i in 1:n_draws) {
  row_idx <- draw_indices[i]
  
    int_phiP     <- outmat[row_idx, "int.phiP"]
    b_sst_P      <- outmat[row_idx, "b.sst.P"] 
    b_sst_1      <- outmat[row_idx, "b.sst.1"]
    b_sst_2.3      <- outmat[row_idx, "b.sst.2.3"]
    b_sst_4.6      <- outmat[row_idx, "b.sst.4.6"]
    b_sst_7.9      <- outmat[row_idx, "b.sst.7.9"]
    b_sst_A      <- outmat[row_idx, "b.sst.A"]
    beta_DD      <- outmat[row_idx, "beta.DD"]
    
    int_phi1     <- outmat[row_idx, "int.phi1"]
    int_phi2.3   <- outmat[row_idx, "int.phi2.3"]
    int_phi4.6   <- outmat[row_idx, "int.phi4.6"]
    int_phi7.9   <- outmat[row_idx, "int.phi7.9"]
    int_phi10.12 <- outmat[row_idx, "int.phi10.12"]
    int_phi13.16 <- outmat[row_idx, "int.phi13.16"]
    int_phiA     <- outmat[row_idx, "int.phiA"]
    fec_low      <- outmat[row_idx, "int.fec_low"]
    fec_high     <- outmat[row_idx, "int.fec_high"]
    
    # Males
    int_phiPM    <- outmat[row_idx, "int.phiPM"]
    int_phi1M    <- outmat[row_idx, "int.phi1M"]
    int_phi2.3M  <- outmat[row_idx, "int.phi2.3M"]
    int_phi4.6M  <- outmat[row_idx, "int.phi4.6M"]
    int_phi7.9M  <- outmat[row_idx, "int.phi7.9M"]
    int_phi10.12M<- outmat[row_idx, "int.phi10.12M"]
    int_phi13.16M<- outmat[row_idx, "int.phi13.16M"]
    int_phiAM    <- outmat[row_idx, "int.phiAM"]

    mu_P  <- qlogis(int_phiP)
    mu_1  <- qlogis(int_phi1)
  mu_2.3 <- qlogis(int_phi2.3)
  mu_4.6 <- qlogis(int_phi4.6)
  mu_7.9 <- qlogis(int_phi7.9)
  mu_10.12 <- qlogis(int_phi10.12)
  mu_13.16 <- qlogis(int_phi13.16)
  mu_A <- qlogis(int_phiA)
   mu_PM <- qlogis(int_phiPM)
  mu_1M <- qlogis(int_phi1M)
  mu_2.3M <- qlogis(int_phi2.3M)
  mu_4.6M <- qlogis(int_phi4.6M)
  mu_7.9M <- qlogis(int_phi7.9M)
  mu_10.12M <- qlogis(int_phi10.12M)
  mu_13.16M <- qlogis(int_phi13.16M)
  mu_AM <- qlogis(int_phiAM)
  
  Nf_vec <- rep(1000, 25)
  Nm_vec <- rep(1000, 20)
  
  for (y in 1:n_years) {
    N_tot <- sum(Nf_vec) + sum(Nm_vec)
    pop_trajectories[i, y] <- N_tot 
    
    current_sst <- sst_vec[y]
    
    # Apply density dependence and the specific year's SST
    phiP_current  <- plogis(mu_P  + (b_sst_P * current_sst) - beta_DD*(N_tot/100000))
    phiPM_current <- plogis(mu_PM + (b_sst_P * current_sst) - beta_DD*(N_tot/100000))
    phi1_current  <- plogis(mu_1  + (b_sst_1 * current_sst)) 
    phi1M_current <- plogis(mu_1M + (b_sst_1 * current_sst)) 
    phi2.3_current  <- plogis(mu_2.3  + (b_sst_2.3 * current_sst)) 
    phi2.3M_current <- plogis(mu_2.3M + (b_sst_2.3M * current_sst)) 
    phi4.6_current  <- plogis(mu_4.6  + (b_sst_4.6 * current_sst)) 
    phi4.6M_current  <- plogis(mu_4.6M  + (b_sst_4.6M * current_sst)) 
    phi7.9_current  <- plogis(mu_7.9  + (b_sst_7.9 * current_sst)) 
    phi7.9M_current  <- plogis(mu_7.9M  + (b_sst_7.9M * current_sst)) 
    phi10.12_current  <- plogis(mu_10.12  + (b_sst_10.12 * current_sst))
    phi10.12M_current  <- plogis(mu_10.12M  + (b_sst_10.12M * current_sst))
    phi13.16_current  <- plogis(mu_13.16  + (b_sst_13.16 * current_sst))
    phi13.16M_current  <- plogis(mu_13.16M  + (b_sst_13.16M * current_sst))
    phiA_current  <- plogis(mu_A  + (b_sst_A * current_sst))
    phiAM_current  <- plogis(mu_AM  + (b_sst_AM * current_sst))

    pups_born <- (sum(Nf_vec[5:10])*fec_low) + (sum(Nf_vec[11:23])*fec_high)
    
    Nm_vec_next <- numeric(25)
    Nf_vec_next <- numeric(20)
    
    Nm_vec_next[1] <- pups_born * 0.5
    Nf_vec_next[1] <- pups_born * 0.5
    
    Nm_vec_next[2] <- Nm_vec[1] * phiPM_current
    Nf_vec_next[2] <- Nf_vec[1] * phiP_current
    
    Nm_vec_next[3] <- Nm_vec[2] * phi1M_current 
    Nf_vec_next[3] <- Nf_vec[2] * phi1_current
    
    Nm_vec_next[4] <- Nm_vec[3] * phi2.3M_current 
    Nm_vec_next[5] <- Nm_vec[4] * phi2.3M_current 
    Nf_vec_next[4] <- Nf_vec[3] * phi2.3_current
    Nf_vec_next[5] <- Nf_vec[4] * phi2.3_current
    
       Nf_vec_next[6] <- Nf_vec[5] * phi4.6_current
    Nf_vec_next[7] <- Nf_vec[6] * phi4.6_current
    Nf_vec_next[8] <- Nf_vec[7] * phi4.6_current
    Nf_vec_next[9] <- Nf_vec[8] * phi7.9_current
    Nf_vec_next[10] <- Nf_vec[9] * phi7.9_current
    Nf_vec_next[11] <- Nf_vec[10] * phi7.9_current
    Nf_vec_next[12] <- Nf_vec[11] * phi10.12_current
    Nf_vec_next[13] <- Nf_vec[12] * phi10.12_current
    Nf_vec_next[14] <- Nf_vec[13] * phi10.12_current
    Nf_vec_next[15] <- Nf_vec[14] * phi13.16_current
    Nf_vec_next[16] <- Nf_vec[15] * phi13.16_current
    Nf_vec_next[17] <- Nf_vec[16] * phi13.16_current
    Nf_vec_next[18] <- Nf_vec[17] * phi13.16_current
    
    Nm_vec_next[6] <- Nm_vec[5] * phi4.6M_current
    Nm_vec_next[7] <- Nm_vec[6] * phi4.6M_current
    Nm_vec_next[8] <- Nm_vec[7] * phi4.6M_current
    Nm_vec_next[9] <- Nm_vec[8] * phi7.9M_current
    Nm_vec_next[10] <- Nm_vec[9] * phi7.9M_current
    Nm_vec_next[11] <- Nm_vec[10] * phi7.9M_current
    Nm_vec_next[12] <- Nm_vec[11] * phi10.12M_current
    Nm_vec_next[13] <- Nm_vec[12] * phi10.12M_current
    Nm_vec_next[14] <- Nm_vec[13] * phi10.12M_current
    Nm_vec_next[15] <- Nm_vec[14] * phi13.16M_current
    Nm_vec_next[16] <- Nm_vec[15] * phi13.16M_current
    Nm_vec_next[17] <- Nm_vec[16] * phi13.16M_current
    Nm_vec_next[18] <- Nm_vec[17] * phi13.16M_current
      
      for (a in 18:24) {
        Nf_vec_next[a+1] <- Nf_vec[a] * phiA_current
      }
    
        for (a in 18:19) {
        Nm_vec_next[a+1] <- Nm_vec[a] * phiAM_current
      }
    
    Nm_vec <- Nm_vec_next
    Nf_vec <- Nf_vec_next
  }
  
  # take mean of the last 50 years to find the equilibrium band
  realized_K[i] <- mean(pop_trajectories[i, (n_years-100):n_years])
    
  ##calculate MNLP
  K_current <- sum(Nf_vec) + sum(Nm_vec)
  prop_f <- Nf_vec/K_current
  prop_m <- Nm_vec/K_current
  
  #create a sequence of 100 test population sizes from 1,000 up to K
  n_test_points <- 100
  N_test_seq <- seq(1000, K_current, length.out = n_test_points)
  net_growth <- numeric(n_test_points)
  
  # define the average environmental condition for the MNPL test
  eval_sst <- mean(sst_vec)
  
  # across test sizes
  for (step in 1:n_test_points) {
    N_test <- N_test_seq[step]
    
    # Scale the test population according to the SAD
    Nf_temp <- prop_f * N_test
    Nm_temp <- prop_m * N_test
    
    #for 1 year of the projection
    # Apply density dependence to each test size
    phiP_temp  <- plogis(mu_P  + (b_sst_P * eval_sst) - beta_DD*(N_test/100000))
    phiPM_temp <- plogis(mu_PM + (b_sst_P * eval_sst) - beta_DD*(N_test/100000))
    
    # Apply average SST to yearlings and juveniles
    phi1_temp  <- plogis(mu_1  + (b_sst_1 * eval_sst)) 
    phi1M_temp <- plogis(mu_1M + (b_sst_1 * eval_sst)) 
    phi2.3_temp  <- plogis(mu_2.3  + (b_sst_2.3 * eval_sst)) 
    phi2.3M_temp <- plogis(mu_2.3M + (b_sst_2.3 * eval_sst)) 
    phi7.9_temp  <- plogis(mu_7.9  + (b_sst_7.9 * eval_sst))
    phi10.12_temp  <- plogis(mu_10.12  + (b_sst_A * eval_sst))
    phi13.16_temp  <- plogis(mu_13.16  + (b_sst_A * eval_sst))
    phiA_temp  <- plogis(mu_A  + (b_sst_A * eval_sst))
    
    # Calculate births
    pups_born <- (sum(Nf_temp[5:10])*fec_low) + (sum(Nf_temp[11:23])*fec_high)
    
    Nm_temp_next <- numeric(25)
    Nf_temp_next <- numeric(25)
    
    Nm_temp_next[1] <- pups_born * 0.5
    Nf_temp_next[1] <- pups_born * 0.5
    
    Nm_temp_next[2] <- Nm_temp[1] * phiPM_temp
    Nf_temp_next[2] <- Nf_temp[1] * phiP_temp
    
    Nm_temp_next[3] <- Nm_temp[2] * phi1M_temp 
    Nf_temp_next[3] <- Nf_temp[2] * phi1_temp
    
    Nm_temp_next[4] <- Nm_temp[3] * phi2.3M_temp 
    Nm_temp_next[5] <- Nm_temp[4] * phi2.3M_temp 
    Nf_temp_next[4] <- Nf_temp[3] * phi2.3_temp
    Nf_temp_next[5] <- Nf_temp[4] * phi2.3_temp
    
    Nf_temp_next[6] <- Nf_temp[5] * int_phi4.6
    Nf_temp_next[7] <- Nf_temp[6] * int_phi4.6
    Nf_temp_next[8] <- Nf_temp[7] * int_phi4.6
    Nf_temp_next[9] <- Nf_temp[8] * phi7.9_temp
    Nf_temp_next[10] <- Nf_temp[9] * phi7.9_temp
    Nf_temp_next[11] <- Nf_temp[10] * phi7.9_temp
    Nf_temp_next[12] <- Nf_temp[11] * phi10.12_temp
    Nf_temp_next[13] <- Nf_temp[12] * phi10.12_temp
    Nf_temp_next[14] <- Nf_temp[13] * phi10.12_temp
    Nf_temp_next[15] <- Nf_temp[14] * phi13.16_temp
    Nf_temp_next[16] <- Nf_temp[15] * phi13.16_temp
    Nf_temp_next[17] <- Nf_temp[16] * phi13.16_temp
    Nf_temp_next[18] <- Nf_temp[17] * phi13.16_temp
    
    Nm_temp_next[6] <- Nm_temp[5] * int_phi4.6M
    Nm_temp_next[7] <- Nm_temp[6] * int_phi4.6M
    Nm_temp_next[8] <- Nm_temp[7] * int_phi4.6M
    Nm_temp_next[9] <- Nm_temp[8] * int_phi7.9M
    Nm_temp_next[10] <- Nm_temp[9] * int_phi7.9M
    Nm_temp_next[11] <- Nm_temp[10] * int_phi7.9M
    Nm_temp_next[12] <- Nm_temp[11] * int_phi10.12M
    Nm_temp_next[13] <- Nm_temp[12] * int_phi10.12M
    Nm_temp_next[14] <- Nm_temp[13] * int_phi10.12M
    Nm_temp_next[15] <- Nm_temp[14] * int_phi13.16M
    Nm_temp_next[16] <- Nm_temp[15] * int_phi13.16M
    Nm_temp_next[17] <- Nm_temp[16] * int_phi13.16M
    Nm_temp_next[18] <- Nm_temp[17] * int_phi13.16M
    
    for (a in 18:24) {
      Nm_temp_next[a+1] <- Nm_temp[a] * int_phiAM
      Nf_temp_next[a+1] <- Nf_temp[a] * phiA_temp
    }
    
    # Calculate net growth for this test size
    N_next_total <- sum(Nm_temp_next) + sum(Nf_temp_next)
    net_growth[step] <- N_next_total - N_test
  }
  
  # identify max productivity level
  max_idx <- which.max(net_growth)
  mnpl_estimates[i] <- N_test_seq[max_idx]
  mnpl_ratio[i] <- mnpl_estimates[i]/K_current
  
}

mean_trajectory <- colMeans(pop_trajectories)

ci_lower_traj <- apply(pop_trajectories, 2, quantile, probs = 0.025)
ci_upper_traj <- apply(pop_trajectories, 2, quantile, probs = 0.975)

# realized K metrics (from the last 100 years of each draw)
realized_K_mean <- mean(realized_K)
real_K_CI <- quantile(realized_K, probs = c(0.025, 0.975), names = F)

mean_mnpl <- mean(mnpl_estimates)
mnpl_CI <- quantile(mnpl_estimates, probs = c(0.025, 0.975), names = F)
mean_mnpl_ratio <- mean(mnpl_ratio)

N_current <- outmat[draw_indices, "N_total[35]"] 
mean_N_current <- mean(N_current)

# MNPL_vector <- realized_K * 0.60  # statutory default
MNPL_vector <- mnpl_estimates       # empirical matrix results
MNPL_mean <- mean(mnpl_estimates)

# probability that the stock is within OSP: how often Ntot[35] is larger than MNPL 
OSP_check <- N_current > MNPL_vector
prob_within_OSP <- mean(OSP_check)

# ratio of Ntot[35] relative to MNPL; less than 1.0 means depleted, above means within OSP
depletion_ratio <- N_current/MNPL_vector
mean_ratio <- mean(depletion_ratio)
ci_ratio <- quantile(depletion_ratio, probs = c(0.025, 0.975), names = F)

# print(paste("Probability stock is within OSP:", round(prob_within_OSP * 100, 1), "%"))
# print(paste("Current abundance is", round(mean_ratio, 2), "times the MNPL."))
# print(paste("95% CI of that ratio:", round(ci_ratio[1], 2), "-", round(ci_ratio[2], 2)))

# print(paste("Realized K:", round(realized_K_mean)))
# print(paste("95% CI:", round(real_K_CI[1]), "-", round(real_K_CI[2])))

# plot(1:n_years, mean_trajectory, type = "n",
#      main = "Population trajectory with oscillating SST (Realized K)",
#      xlab = "Projection year", ylab = "Total abundance",
#      ylim = c(0, max(ci_upper_traj) * 1.05)) # Give a 5% buffer at the top
# 
# polygon(c(1:n_years, rev(1:n_years)),
#         c(ci_lower_traj, rev(ci_upper_traj)),
#         col = rgb(0, 0, 1, alpha = 0.15), border = NA)
# 
# lines(1:n_years, mean_trajectory, col = "blue", lwd = 2)
# 
# abline(h = realized_K_mean, col = "darkred", lty = 2, lwd = 2)
# abline(h = real_K_CI[1], col = "darkred", lty = 3, lwd = 1.5)
# abline(h = real_K_CI[2], col = "darkred", lty = 3, lwd = 1.5)
# 
# legend("bottomright",
#        legend = c("Mean trajectory", "95% CI", "Mean realized K", "95% CI"),
#        col = c("blue", rgb(0, 0, 1, alpha = 0.15), "darkred", "darkred"),
#        lty = c(1, 1, 2, 3),
#        lwd = c(2, 10, 2, 1.5),
#        bty = "n")

#plot productivity curve
# peak_growth <- max(net_growth)
# peak_N <- N_test_seq[which.max(net_growth)]
# 
# plot(N_test_seq, net_growth, type = "l", col = "forestgreen", lwd = 3,
#      main = "Density-Dependent productivity curve",
#      xlab = "Total population size (N)", 
#      ylab = "Net annual growth (new individuals)",
#      xlim = c(0, max(N_test_seq) * 1.05))
# 
# # add a horizontal line at 0 where growth stops
# abline(h = 0, col = "black", lty = 1, lwd = 1)
# 
# # add a vertical line at K where the curve hits 0
# abline(v = K_current, col = "blue", lty = 2, lwd = 2)
# 
# # add a vertical line at MNPL - the peak of the curve
# abline(v = peak_N, col = "red", lty = 2, lwd = 2)
# 
# points(peak_N, peak_growth, pch = 16, col = "red", cex = 1.5)
# 
# legend("bottomleft", 
#        legend = c("Annual productivity", "MNPL (peak growth)", "Carrying capacity (K)"),
#        col = c("forestgreen", "red", "blue"),
#        lty = c(1, 2, 2), lwd = c(3, 2, 2), pch = c(NA, NA, NA), bty = "n")

```


#### CSL population dynamics and environmental effects  

*Project status*  

- MR model estimates look great  
-- Haven't incorporated dead stranded IDs or lepto/DA  
- Prototype IPM model estimates look like they're working (model takes a *long* long time to run)  
- Estimates of management metrics between IPM and MR models align - this is also great and not to be taken for granted   
- Methods drafted, results visualized  
- Medium to-do.... think a bit about the density dependence mechanism in the IPM given the estimated effects of ENSO on not just pups... yearlings and juveniles have a harder time finding food too? 

*Draft fodder for future manuscript effort...* 

#### Introduction 

CSL as indicator species in a dynamic CCE 
- Importance of monitoring sentinel species to understand population dynamics, provide updated management and conservation metrics, and understand responses of a species sensitive to environmental variability in an ecosystem notorious for cyclic phases  
- Climate impacts the food web and health  
- Opportunities afforded by a well-studied (and observable) species in a dynamic environment  

Estimating density dependence for long-lived top predator  
- Challenges  
- Different approaches  

CSL historical and contemporary context  
- ups and downs and connections with ENSO  
- prominent species along the coast  
- foraging ecology/fisheries interactions  
- human impacts and disease/stranding/public concern  

Current importance  
-- estimates growing outdated  
-- improvements upon previous estimation, and more SST evidence  

Aims: upgrade Laake et al. and provide management metrics    
- Laake et al implemented a two-step approach: survival estimates fed into population projection model and then separately derive K and MNPL  
-- Benefits: computationally efficient  
-- Disadvantages: (1) does not fully integrate uncertainty in MR estimates within the population projection model and therefore does not resolve discrepancies between the two datasets based on relative precision of the data streams; (2) DD not estimated with respect to a particular vital rate (doesn't specify mechanism of dependence) and assumes downturns in abundance are due to density instead of environmental effect  
- Fully integrated IPM addresses these issues: (1) weights uncertainty in each data stream to settle discrepancies; uncertainty from all components propagated into final posteriors of *K* and MNPL, and (2) estimates DD with respect to an ecological mechanism and facilitates specifying how transient dynamics (such as small cohort of pups from ENSO event) propagates through the age structure over the next decade   
- Full IPM distinguishes between ENSO noise and density signal - what would *K* be in an average environment -- may be a higher/truer carrying capacity instead of logistic growth model where a lot of high SST years could look like approaching *K*   
- In the IPM, MNPL and *K* are emergent properties derived from age-structured vital rates. Laake et al. used the $\theta$ parameter to force the productivity curve into a specific mathematical shape, which could skew MNPL relative to *K* (generally 0.6 for marine mammals instead of 0.5K that would be represented by perfect symmetry) whereas the IPM will organically generate right-skewed productivity curve based on vital rates  
- Ultimately, this model enhances ecological realism  

<!-- Literature: -->
<!-- - Hammill et al. 2021 for northwest atlantic grey seals; similar density dependence framework -->
<!-- - Ersalman et al. 2024-2025 on baltic ringed seals? env focus on younglings -->
<!-- - Tinker et al. 2023 for NW atlantic harp seals; K and ice-related pup mortality -->

<!-- Hammill et al. and Tinker et al still use mathematical shape parameter theta -->
 
#### Methods

##### Study system   

##### Data collection and curation  

- Marked cohort sizes and years (every year? similar approximate samples sizes?)
-- approximately 500 marked annually (!!wow!!) 1990-2005, 300-400 from 2006-2015, 250-300 thereafter.

<!-- (table(ch_dat_m$cohort) + table(ch_dat_f$cohort)) -->

- Pup counts and imputation to fill any gaps.

- SST data and seasonality as relevant to pups vs. non-pups.


##### Statistical analyses

We developed a fully integrated Bayesian population model to jointly estimate vital rates and population dynamics for California sea lions using data collected from the California Coast from 1990-2025. The IPM framework simultaneously evaluates a Cormack-Jolly-Seber (CJS; citation) mark-recapture model and a state-space demographic projection matrix, linked by shared age- and sex-specific survival parameters. To resolve latent fecundity and anchor population trajectories, we incorporated annual pup counts treated as a census without observation error.

*Vital rates subcomponent model*

We modeled the individual capture histories using a CJS framework to estimate age- and sex-specific apparent survival ($\phi$) and detection probabilities ($p$). The state process was modeled using a categorical distribution tracking individuals across distinct biological age classes, up to a maximum modeled lifespan of 24 years.

Survival probabilities were estimated for pups, ages 1, 2-3, 4-6, 7-9, 10-12, 13-16, and 17+ for both males and females. Survival probabilities were estimated on the logit scale and modeled as a function of baseline age- and sex-specific intercepts ($\mu$) and environmental stochasticity ($\beta_{\text{SST}}$). Additional temporal variation not attributable to the effects of SST were not included and pup birth weight, while known to be influential in survival outcomes, was not included as a covariate here due to its correlation with annual SST (i.e., pup weights negatively correlated with SST; citation or create additional supplemental figure) . Because density dependence in pinnipeds acts most strongly on early life stages (i.e., pups), so the density effect ($\beta_{\text{DD}}$) was isolated to pup survival ($\phi_P$). Total annual population size, $N_t$, was the summation of all male and female age classes and scaled by a constant ($100,000$) to facilitate Markov chain Monte Carlo (MCMC) convergence. For an individual $i$ at time $t$ and sex $s$, pup survival was parameterized as:

$$\text{logit}(\phi_{p, i, t}) = \mu_p + \beta_{\text{SST}_p} \cdot \text{SST}_{t} - \beta_{\text{DD}} \cdot \left( \frac{N_{t}}{100,000} \right)$$

Juvenile and adult survival and fecundity were assumed to be density-independent and driven solely by baseline rates and environmental stochasticity. Thus, survival for older age classes (yearlings through adults) was parameterized with age-specific intercepts and distinct sea surface temperature values ($\text{SST}_{t}$) relevant to foraging ecology:

$$\text{logit}(\phi_{a, i, t}) = \mu_a + \beta_{\text{SST}, a} \cdot \text{SST}_{t}$$
Detection probailities ($p$) were assumed to be constant between sexes and estimated for each age class with temporal variance pooled across ages estimated using a shrinkage prior, but detection was assumed to be constant across sexes, natal regions, and resight location. 

In the mark-resight subcomponent model, we assumed that marks were not lost, misread, or missed when an animal is encountered, and that all individuals within a specific age and sex class share the same underlying probabilities of survival and detection in a given year.

*Population projection subcomponent model*

To evaluate population-level dynamics, we constructed a sex-specific deterministic and stochastic projection model using survival rates estimated as above. We assumed a post-breeding census timeline, meaning the annual pup counts represent reproduction by females that successfully survived from year $t$ into year $t+1$.

Because pup counts were treated as a census, the number of pups at age 0 ($N_{1, t}$) was assigned deterministically from field observations, assuming a 50:50 sex ratio. To avoid MCMC initialization failures, the latent age classes for the first year of the time series were initialized using a stable age distribution derived from baseline survival probability approximations.

Population structure and age-specific abundance for each age class $a$ were modeled via Poisson distributions (to facilitate MCMC sampling instead of more constrained binomial draws) using the predicted annual age-specific survival probabilities (${\phi}_{a, t}$) accounting for changes in SST over the study period multiplied by the age-specific abundance in the previous year:

$$N_{a+1, t+1} \sim \text{Poisson}({\phi}_{a, t}\cdot N_{a, t})$$

Individuals reaching the terminal age class (age 24) were assumed to undergo complete reproductive and survival senescence and were removed from the matrix in subsequent time steps.

*Integrated population model*

To link the unobserved adult female population to the perfectly observed pup census, we modeled reproduction using a post-breeding census assumption. Pups born in year $t+1$ were calculated as a function of the females that successfully survived into year $t+1$.

Fecundity was estimated as a latent parameter divided into two biological tiers: a lower fecundity rate for younger reproductive females ($F_{\text{4:9}}$, Ages 4–9) and a higher fecundity rate for prime-age adults ($F_{\text{10:22}}$, Ages 10–22). Assuming an equal sex ratio at birth, the expected total number of pups produced was calculated as:

$$E[Pups_{t+1}] = \left( \sum_{a=5}^{10} N_{f, a, t+1} \cdot F_{\text{4:9}} \right) + \left( \sum_{a=11}^{23} N_{f, a, t+1} \cdot F_{\text{10:22}} \right)$$

The final integration of the demography and the count data was achieved by evaluating the expected pup production as estimated in the population growth equations in terms of the census data ($y_{t+1}$) using a Poisson likelihood constraint:

$$y_{t+1} \sim \text{Poisson}(E[Pups_{t+1}])$$

Any divergence between expected reproduction and the census data directly penalizes the likelihood of the adult female survival and fecundity estimates, ensuring mathematical tension and precise demographic alignment across the integrated framework.

Assuming the mark-recapture data ($\mathbf{y}_{MR}$) and the pup census counts ($\mathbf{y}_{count}$) are conditionally independent, the joint likelihood of the model ($L_{IPM}$) is the product of their separate likelihoods:

$$L_{IPM}(\theta|\mathbf{y}_{MR},\mathbf{y}_{count})=L_{MR}(\theta|\mathbf{y}_{MR})\times L_{count}(\theta|\mathbf{y}_{count})$$
where $\theta$ represents the joint vector of all estimated parameters (survival, detection, fecundity, and environmental/density coefficients). By evaluating this joint likelihood, the model allows the high-precision pup census data to inform and constrain the latent vital rates that cannot be fully resolved by the mark-recapture data alone.

*Estimating theoretical and realized carrying capacity:*

In highly variable marine ecosystems, traditional phenomenological population models often confound climate-driven mortality with density-dependent limitation. Because legacy models for this stock estimated carrying capacity (*K*) by fitting a generalized logistic growth curve directly to reconstructed abundance time series, sudden population crashes driven by marine heatwaves (e.g., El Niño events) were mathematically interpreted as the population reaching its density-dependent ceiling. This approach likely underestimates the true biological potential of the stock. To resolve this, we utilized the mechanistic framework of the IPM, which simultaneously estimates the effects of population density and environmental variability (e.g., SST) on age-specific survival. By decoupling these factors, we defined and estimated two distinct carrying capacity metrics: theoretical *K* and realized *K*.

Because density dependence was parameterized mechanistically as a penalty on pup survival rather than through an overarching population growth curve, carrying capacity ($K$) was not estimated as an explicit parameter in the joint likelihood. Instead, $K$ was treated as an emergent property of the estimated vital rates. For each posterior draw, $K$ was derived numerically via a post-hoc simulation. We developed a deterministic, sex-specific matrix projection model parameterized directly from the joint posterior distribution of the IPM. The matrix tracked 25 age classes for both males and females. Annual pup production was calculated using age-specific fecundity rates applied to females in age classes 5 through 23, assuming a 50:50 sex ratio at birth.

To propagate parameter uncertainty, we randomly sampled 500 MCMC draws from the posterior distribution. For each draw, the matrix was initialized with a dummy seed population and projected forward for 400 years to ensure the population had ample time to reach equilibrium. Survival for pups, yearlings, and juveniles in the projection was dynamically updated at each time step based on the respective SST and density-dependence ($\beta_{DD}$) coefficients estimated by the IPM.

<!-- As the simulated population grew, the density-dependent penalty ($\beta_{DD}$) incrementally suppressed pup survival. $K$ was defined as the equilibrium population size at which the number of deaths perfectly balanced the number of births (population growth rate $\lambda=1$): -->

$$K=\lim_{t\to\infty}\sum_{a=1}^{25}N_{a,t}\quad\text{where}\quad N_{t+1}=N_{t}$$


Theoretical Carrying Capacity:
Theoretical *K* represents the absolute biological ceiling of the ecosystem based purely on space and baseline resources, absent of environmental extremes. To estimate Theoretical *K*, we projected the population forward holding the SST covariate constant at 0 (representing the mean-centered baseline temperature). Under this scenario, population growth is limited exclusively by the density-dependent penalty on pup survival. Theoretical *K* for each posterior draw was defined as the stabilized total abundance at year 400.

Realized Carrying Capacity:  
Because the California Current Ecosystem is characterized by frequent, cyclical marine heatwaves that acutely depress survival across multiple juvenile age classes, Theoretical *K* is ecologically unrealistic. Therefore, we estimated Realized *K*, defined as the long-term average abundance the population fluctuates around when subjected to historical frequencies of environmental variance. To estimate Realized *K*, we subjected the simulated population to a dynamic thermal regime rather than a constant baseline. At each annual time step of the 400-year projection, an SST anomaly was drawn from a bootstrapped distribution of historical SST anomalies from 1990-2024. Realized *K* for each posterior draw was calculated as the mean total abundance over the final 100 years of the projection, capturing the dynamic equilibrium band. 

*Estimation of MNPL and OSP Status*  

To evaluate the population status relative to the Marine Mammal Protection Act (MMPA) management metrics, we calculated the Maximum Net Productivity Level (MNPL) and Optimal Sustainable Population (OSP) status from the posterior draws of the $K$ simulations. 

Rather than assuming a rigid generalized logistic (theta-logistic) shape parameter, MNPL was derived numerically by tracking the absolute population growth ($\Delta N$) at each time step of the *post-hoc* simulation. To propagate the uncertainty from our IPM posterior, MNPL was calculated iteration-by-iteration as $0.60 \times \text{Realized } K_i$. MNPL was identified as the population size where net production (the number of individuals added to the population annually) reached its global maximum:

$$MNPL=N_{t}\quad\text{where}\quad\max(N_{t+1}-N_{t})\text{ occurs}$$

To determine the stock's status relative to OSP, we compared the terminal year abundance estimate from the IPM ($N_{2024}$) to the calculated MNPL. A population is considered to be within its OSP if this ratio is greater than or equal to 1:

$$OSP_{status}=\frac{N_{current}}{MNPL}$$

This numeric derivation ensures that the management metrics are driven entirely by the species' empirical, age-structured vital rates and specific density-dependent survival thresholds estimated by the IPM.

##### Implementation and diagnostics 

This Bayesian state-space integrated population model was implemented in NIMBLE (NIMBLE Development Team 2019) using three Markov chain Monte Carlo (MCMC) chains with 30,000 iterations, 10,000 burn-in, and a thinning rate of 3 within the R programming environment (R Core Team 2020). Convergence was assessed by examining MCMC chains and a Brooks-Gelman-Rubin statistic (Gelman & Rubin 1992; Brooks & Roberts 1998) < 1.1. We summarized the results of the full model according to the proportion of MCMC chain samples for the environmental effects that was above versus below zero. 

#### Results  

Results described from the IPM are still only a subset (50-100 marked individuals per year) of the full dataset. 

##### Vital rates  

```{r phi plot, eval = T}

phi_plot
phi_plot_compare

```

The first plot shows sex- and age class-specific survival probabilities and associated 95% credible intervals. The second plot shows very strong agreement between these MR-only model estimates and those from the IPM for female survival probabilities. 


```{r p plot, eval = T}

p_plot

```

Sex- and age-specific annual detection probabilities, though yearlings seem low? Should it be sex-specific for yearlings and juveniles too? Males lower makes sense too?  

```{r fec plot, eval = T}

fec_plot

```

Fecundity estimates for younger and older adult females based on the integrated population model. 

##### Covariate effects  

```{r cov plots, eval = T}

sst_eff_plot
sst_preds_plot
enso_eff_plot

```

Figure (A): logit-scale (positive versus negative) effect of SST for each age shows that pups and yearlings are largely affected similarly and that there are negative effects estimated for all age classes except 4-6yo's. Plain language interpretation: survival is lower for all age classes (except 4-6) in years when SST is high and the effect is strongest for pups and yearlings. *Note to self - recheck whether these are sex-specific effects?*

Figure (B): probability-scale effects of the SST covariate effect for each age: survival goes from approximately 75% in years with lower SST to approximately 30% for pups and yearlings when SST is high. 

Figure (C): I'm calling this the "ENSO memory" effect: whether individuals are affected in subsequent years if born in an ENSO year. This shows that (for example) survival is high at ages 2-6 for individuals born in ENSO years and lower for adults that were born in ENSO years. We should talk through this a bit more, but my guess is that this is more showing us a cyclic effect rather than some sort of lasting ENSO effect that pups carry with them as they age. Once pups that were born in an ENSO event have reached ages 2-6, those years are probably cold, and once those same pups that were born in an ENSO event have reached 7+ survival is maybe lower again because they're back into an ENSO event. 

##### Management metrics: *K* and MNPL 

*Mark-recapture reconstruction estimates* 

The two-step approach (e.g., Laake et al update) produces the following estimates of *K* and MNLP.... keeping in mind that these are lower because the model can't distinguish between density effects versus environmental effects (e.g., downturns in ENSO years can be misconstrued as density dependence). 

Carrying capacity: `r format(round(summary_table[1,1]), big.mark = ',')` (95% CI: `r format(round(summary_table[1,2]), big.mark = ',')`-`r format(round(summary_table[1,3]), big.mark = ',')`).

MNPL: `r format(round(summary_table[1,4]), big.mark = ',')` (95% CI: `r format(round(summary_table[1,5]), big.mark = ',')`-`r format(round(summary_table[1,6]), big.mark = ',')`).

```{r post hoc MR metrics 1, eval = T}

# print(round(summary_table))

plot(x = 1990:2024, y = recon$Ntot, type = "l", col = "darkblue", lwd = 2, 
     main = "Reconstructed population trajectory",
     ylab = "Total abundance", xlab = "")

```

Reconstructed population trajectory based on mark-recapture estimates (does this look reasonable?). 


```{r post hoc MR metrics 2, eval = T}

# print(round(summary_table))

plot(x = 1990:2024, y = implied_fecundity, type = "b", col = "darkred", pch = 16,
     # main = "Implied birth rate",
     ylab = "Fecundity (pups per adult female)", xlab = "")

```

Implied fecundity rate derived from reconstructed population (is this about right for CSL? Seems a bit low, or at least a bit volatile). Could consider two fecundity rates as in the IPM. 
 
*Integrated population model estimates*   

**These are just prototypes!!**

The estimate of final-year abundance: `r format(round(mean_N_current), big.mark = ',')`. 

Estimate of theoretical *K*: `r format(round(theo_K), big.mark = ',')` (95% CI: `r format(round(theo_K_CI), big.mark = ',')`).  

Estimate of realized *K*: `r format(round(realized_K_mean), big.mark = ',')`. 

The estimate of MNPL is `r format(round(mean_mnpl), big.mark = ',')` (95% CI: `r format(round(mnpl_CI), big.mark = ',')`) and current abundance is `r format(round(mean_ratio, 2), big.mark = ',')` times the MNPL. 


``` {r IPM metrics 2, eval = T}

#theoretical K
# theo_K
# theo_K_CI

#realized K
# print(paste('Realized K:', round(realized_K_mean)))
# print(paste("95% CI:", round(real_K_CI[1]), "-", round(real_K_CI[2])))

#current abundance
# mean_N_current #<- mean(N_current)

#MNPL
# mean_mnpl #<- mean(mnpl_estimates)
# mean_mnpl_ratio #<- mean(mnpl_ratio)

# probability that the stock is within OSP: how often Ntot[35] is larger than MNPL 
# prob_within_OSP #<- mean(OSP_check)

# ratio of Ntot[35] relative to MNPL; less than 1.0 means depleted, above means within OSP
# depletion_ratio <- N_current/MNPL_vector
# mean_ratio <- mean(depletion_ratio)
# ci_ratio <- quantile(depletion_ratio, probs = c(0.025, 0.975), names = F)


# print(paste("Probability stock is within OSP:", round(prob_within_OSP * 100, 1), "%"))
# print(paste("Current abundance is", round(mean_ratio, 2), "times the MNPL."))
# print(paste("95% CI of that ratio:", round(ci_ratio[1], 2), "-", round(ci_ratio[2], 2)))

plot(1:n_years, mean_trajectory, type = "n",
     main = "Population trajectory with oscillating SST (Realized K)",
     xlab = "Projection year", ylab = "Total abundance",
     ylim = c(0, max(ci_upper_traj) * 1.05)) # Give a 5% buffer at the top

polygon(c(1:n_years, rev(1:n_years)),
        c(ci_lower_traj, rev(ci_upper_traj)),
        col = rgb(0, 0, 1, alpha = 0.15), border = NA)

lines(1:n_years, mean_trajectory, col = "blue", lwd = 2)

abline(h = realized_K_mean, col = "darkred", lty = 2, lwd = 2)
abline(h = real_K_CI[1], col = "darkred", lty = 3, lwd = 1.5)
abline(h = real_K_CI[2], col = "darkred", lty = 3, lwd = 1.5)

legend("bottomright",
       legend = c("Mean trajectory", "95% CI", "Mean realized K", "95% CI"),
       col = c("blue", rgb(0, 0, 1, alpha = 0.15), "darkred", "darkred"),
       lty = c(1, 1, 2, 3),
       lwd = c(2, 10, 2, 1.5),
       bty = "n")


```


```{r IPM metrics 1, eval = T}

#density dependence
DD_plot

```


#### Discussion

<!-- Laake et al. 2018 metrics:  -->

<!-- - ~275,000 (K) -->
<!-- - MNPL: 183,481... though just based on pups, right...?  -->
<!-- - abundance in 2014: 258k animals, 40% above MNPL so within OSP range -->
<!-- - N/K DD: -1.68 (-3.78,-0.03) -->


#### Appendix

Describe model diagnostics and fit. 

```{r ppc and posterior prior overlaps, eval = T}

pups_ppc_plot


```

Posterior predictive check for number of pups estimated in the IPM relative to census data. Are these low pup counts truly real? 

