---
title: "CSL empirical estimates - spring 2025"
output: html_document
---

```{r setup, include=FALSE, fig.align = 'center'}

knitr::opts_chunk$set(echo = FALSE, eval = T, message = F, warning = F, fig.align = 'center')

library(here) #v1.0.1
library(dplyr) #v1.0.5
library(nimble) #v0.12.1
library(coda)
library(ggh4x)
library(readr)
library(knitr)
library(Hmisc)
library(stringr)

source(here::here('scripts', 'PlotTheme.R'))

```

### Empirical demographic and detection rate estimates (eDPS)  

Survival rates were estimated for ages P, 1, 2, juveniles (ages 3-4) adults (ages 5+), with fixed effects estimating the effect of pup mass and the differences between sex and natal region on survival over 19 years of data. The sample sizes for natal region were imbalanced, with more samples from Pyramid Rock (*n* = 1395) than South Seal Rock (*n* = 290). Fixed effects of sex were estimated for each age group while those of natal region were estimated for pups, yearlings, shared for juveniles (ages 3-4) and adults (ages 5+). Temporal variance was estimated using a random effect of year with a penalized complexity prior, with the random effect estimated for pups, yearlings, and pooled for juveniles (ages 3-4) and adults (ages 5+) jointly for both sex and natal region (i.e., trends were not estimated for each sex nor region but for age groups in the eDPS as a whole). Detection rates were estimated for each age as above with temporal variance constructed in the same fashion, but detection was assumed to be constant across sexes, natal regions, and resight location.  

#### null model
```{r null model}

out <- readRDS(here::here('results', 'out_null.RDS'))
out <- readRDS(here::here('results', 'out_RE.RDS'))
out <- readRDS(here::here('results', 'out_sst.RDS'))

# all_pars <- colnames(out[[1]])
outmat <- as.matrix(out)

post_sum <- data.frame(
  med = apply(outmat, 2, function(x) quantile(x, probs = 0.5, na.rm = T, names = F)),
  lower = apply(outmat, 2, function(x) quantile(x, probs = 0.025, na.rm = T, names = F)),
  upper = apply(outmat, 2, function(x) quantile(x, probs = 0.975, na.rm = T, names = F)))
post_sum$variable <- row.names(post_sum)

#mean rates
# sex.pars <- all_pars[which(grepl('b.sex', all_pars))]
mu.pars <- all_pars[which(grepl('mu.', all_pars))]

b.mass_sum <- post_sum %>% filter(grepl('b.mass', variable)) %>%
  # transform(var_name = ifelse(grepl('b.mass.f', variable), '1', 
  #                                    ifelse(grepl('b.mass.m', variable), '1M',
                  transform(var_name = ifelse(grepl('b.mass1', variable), '1', 
                       ifelse(grepl('b.massP', variable), 'P', NA))) %>%
  transform(sex = ifelse(grepl('M', variable), 'M', 'F')) %>%
  transform(var_name = factor(var_name, levels = c('P', '1'))) 
  
ggplot(b.mass_sum,
       aes(x = var_name, y = med, col = sex, group = sex)) +
  geom_errorbar(aes(x = var_name, ymin=lower, ymax=upper),
                position = position_dodge(0.5),
                width = 0.25, show.legend = F) +
  geom_point(size = 0.8, position = position_dodge(0.5)) +
  geom_hline(aes(yintercept = 0), linetype = 'dashed') +
  xlab('Age class') + ylab(expression(paste('Logit-scale effect of handling mass'))) +
  ggtitle('') +
  plot_theme(legend.position = 'top', panel.border = element_rect(fill = NA),
             plot.title = element_text(hjust = 0.5)) +
  scale_color_manual(values = rainbow2[c(2,5,6,7,8)], name = 'Sex')

#sst
b.sst_sum <- post_sum %>% filter(grepl('b.sst', variable)) %>%
  # transform(var_name = ifelse(grepl('b.mass.f', variable), '1', 
  #                                    ifelse(grepl('b.mass.m', variable), '1M',
                  transform(var_name = gsub('b.sst.', '', variable)) %>%
  transform(var_name = factor(var_name, levels = c('P', '1', '2.3', '4.6', '7.9', 'A'))) 
  
ggplot(b.sst_sum %>% filter(var_name %in% c('P', '1', '2.3', '4.6')),
       aes(x = var_name, y = med)) +
  geom_errorbar(aes(x = var_name, ymin=lower, ymax=upper),
                position = position_dodge(0.5),
                width = 0.25, show.legend = F) +
  geom_point(size = 0.8, position = position_dodge(0.5)) +
  geom_hline(aes(yintercept = 0), linetype = 'dashed') +
  xlab('Age class') + ylab(expression(paste('Logit-scale effect of SST'))) +
  ggtitle('') +
  pres_theme(legend.position = 'top', #panel.border = element_rect(fill = NA),
             plot.title = element_text(hjust = 0.5)) #+
  # scale_color_manual(values = rainbow2[c(2,5,6,7,8)], name = 'Sex')

ggplot(b.sst_sum, aes(x = var_name, y = med)) +
  geom_errorbar(aes(x = var_name, ymin=lower, ymax=upper),
                position = position_dodge(0.5),
                width = 0.25, show.legend = F) +
  geom_point(size = 0.8, position = position_dodge(0.5)) +
  geom_hline(aes(yintercept = 0), linetype = 'dashed') +
  xlab('Age class') + ylab(expression(paste('Logit-scale effect of SST'))) +
  ggtitle('') +
  pres_theme(legend.position = 'top', #panel.border = element_rect(fill = NA),
             plot.title = element_text(hjust = 0.5))


#intercept
#survival
int_dat <- post_sum %>% filter(grepl('int.phi', variable)) %>%
  # bind_rows(post_sum %>% filter(grepl('int.psi', variable))) %>%
  transform(variable = gsub('int.', '', variable)) %>%
  transform(sex = ifelse(grepl('M', variable), 'M', 'F')) %>%
  transform(age = gsub('phi', '', variable)) %>%
  transform(age = gsub('M', '', age)) %>%
   transform(age = factor(age, levels = c('P', 'PM', '1', 
                                                   '1M', '2.3', '2.3M',
                            '4.6', '4.6M', '7.9', '7.9M', '10.12', '10.12M',
                            '13.16', '13.16M', 'A', 'AM'))) 
  
  
ggplot(int_dat %>% filter(grepl('phi', variable)),
       aes(age, med, color = sex)) +
  geom_errorbar(aes(x = age, ymin=lower, ymax=upper),
                position = position_dodge(0.5),
                width = 0.5, show.legend = F) +
  geom_point(size = 0.8, position = position_dodge(0.5)) +
  xlab('Age class') + ylab(expression(paste('Survival probability ',  '(', phi, ')'))) +
  ggtitle('') +
  # facet_wrap(~sex, scales = 'free_x', drop = T) +
  plot_theme(legend.position = 'top', 
             panel.border = element_rect(fill = NA),
             plot.title = element_text(hjust = 0.5)) +
  scale_color_manual(values = rainbow2[c(2,5,6,7,8)], name = 'Sex')

#detection
int_dat <- post_sum %>% filter(grepl('int.p', variable)) %>%
  # bind_rows(post_sum %>% filter(grepl('int.psi', variable))) %>%
  transform(variable = gsub('int.', '', variable)) %>%
  filter(!grepl('phi', variable)) %>%
    transform(sex = ifelse(grepl('M', variable), 'M', 'F')) %>%
  transform(age = gsub('p', '', variable)) %>%
  transform(age = gsub('M', '', age)) %>%
    transform(age = factor(age, levels = c('1', '2.3', '4.6',  
                                                   '7.12', 'A'))) 

ggplot(int_dat,
       aes(age, med, color = sex)) +
  geom_errorbar(aes(x = age, ymin=lower, ymax=upper),
                position = position_dodge(0.5),
                width = 0.5, show.legend = F) +
  geom_point(size = 0.8, position = position_dodge(0.5)) +
  xlab('') + ylab(expression(paste('Detection probability ',  '(', p, ')'))) +
  ggtitle('') +
  # facet_wrap(~sex, scales = 'free_x', drop = T) +
  plot_theme(legend.position = 'top', 
             panel.border = element_rect(fill = NA),
             plot.title = element_text(hjust = 0.5)) +
  scale_color_manual(values = rainbow2[c(2,5,6,7,8)], name = 'Sex')


```


#### Survival over time

```{r temporal variance}


#temporal variance
phi.vals <- post_sum %>%
  filter(grepl('.prob', variable)) %>%
    transform(year = as.numeric(as.character(gsub('.+\\[([0-9]+)\\].*$', '\\1',
                                                variable)))) %>%
  #use when dataframe for phi.prob rather than vector
  # transform(year = as.numeric(as.character(gsub('.+\\[([0-9]+), ([0-9]+)\\].*$', '\\2', 
  #                                               variable)))) %>% 
  # transform(region = gsub('.+\\[([0-9]+), ([0-9]+)\\].*$', '\\1', variable)) %>%
  transform(variable = sub("\\[.*", "", variable)) %>%
  transform(rate = ifelse(grepl('phi', variable), 'Survival', 'Detection')) %>%
  transform(age = gsub('.prob', '', variable)) %>%
  transform(age = ifelse(rate == 'Survival', gsub('phi', '', age), 
                         ifelse(rate == 'Detection', gsub('p', '', age), NA))) %>%
  transform(sex = ifelse(grepl('M', age), 'Male', 'Female')) %>%
  transform(age = gsub('M', '', age)) %>%
  transform(age = factor(age, levels = c('P', '1', '2.3', '4.6', '7.9', '10.12', '13.16', '7.12', 'A'))) %>%
  # merge(mean.phi.vals %>% filter(site == 'Pyramid Rock'), by = c('age'), suffixes = c('', '.mean'))
  transform(exclude_det = ifelse(age == '2.3' & year < 2, 'Y',
                                   ifelse(age == '4.6' & year < 4, 'Y',
                                    ifelse(age == '7.12' & year < 7, 'Y',
                                           ifelse(age == 'A' & year < 13, 'Y', 'N'))))) %>%
  transform(exclude_phi = ifelse(age == '1' & year < 2, 'Y',
                             ifelse(age == '2.3' & year < 3, 'Y',
                                   ifelse(age == '4.6' & year < 5, 'Y',
                                    ifelse(age == '7.9' & year < 7, 'Y',
                                           ifelse(age == '10.12' & year < 10, 'Y',
                                                  ifelse(age == '13.16' & year < 13, 'Y',
                                           ifelse(age == 'A' & year < 16, 'Y', 'N'))))))))

#temporal variance with mean rates; intercept represents female, so can't show males
ggplot(phi.vals %>% filter(rate == 'Detection' & exclude_det != 'Y' & year != 20),
       aes(year, med)) +
  geom_errorbar(aes(x = year, ymin=lower, ymax=upper), position = position_dodge(0.5),
                width = 0.5, show.legend = F) +
  geom_point(size = 0.8, position = position_dodge(0.5)) +
  geom_line(linewidth = 0.7, position = position_dodge(0.5)) +
  # geom_line(aes(year, med.mean), col = 'grey20', linetype = 'dotted') +
  xlab('') + ylab(expression(paste('Detection probability ',  '(', p, ')'))) +
  ggtitle('') +
    facet_grid(age ~ sex) +
  plot_theme(legend.position = 'top', panel.border = element_rect(fill = NA),
             plot.title = element_text(hjust = 0.5)) +
  guides(color = guide_legend("", nrow = 1, byrow = T)) +
  scale_color_manual(values = rainbow2[c(2,3,5,6,7,8)], name = '') +
  scale_x_continuous(breaks = c(seq(1, 25, by = 5)), 
                     labels = c(seq(2001, (2001+n_occasions-1), by = 5)))

#survival
ggplot(phi.vals %>% filter(rate == 'Survival' & exclude_phi != 'Y'),
       aes(year, med)) +
  geom_errorbar(aes(x = year, ymin=lower, ymax=upper), position = position_dodge(0.5),
                width = 0.5, show.legend = F) +
  geom_point(size = 0.8, position = position_dodge(0.5)) +
  geom_line(linewidth = 0.7, position = position_dodge(0.5)) +
  # geom_line(aes(year, med.mean), col = 'grey20', linetype = 'dotted') +
  xlab('') + ylab(expression(paste('Survival probability ',  '(', phi, ')'))) +
  ggtitle('') +
    facet_grid(age ~ sex) +
  plot_theme(legend.position = 'top', panel.border = element_rect(fill = NA),
             plot.title = element_text(hjust = 0.5)) +
  guides(color = guide_legend("", nrow = 1, byrow = T)) +
  scale_color_manual(values = rainbow2[c(2,3,5,6,7,8)], name = '') +
  scale_x_continuous(breaks = c(seq(1, 25, by = 5)), labels = c(seq(2000, (2000+n_occasions-1), by = 5)))

ggplot(phi.vals %>% filter(rate == 'Survival' & exclude_phi != 'Y' &
                             age %in% c('P', '1') & sex == 'Female'),
       aes(year, med)) +
  geom_errorbar(aes(x = year, ymin=lower, ymax=upper), position = position_dodge(0.5),
                width = 0.5, show.legend = F) +
  geom_point(size = 0.8, position = position_dodge(0.5)) +
  geom_line(linewidth = 0.7, position = position_dodge(0.5)) +
  # geom_line(aes(year, med.mean), col = 'grey20', linetype = 'dotted') +
  xlab('') + ylab(expression(paste('Survival probability ',  ' (', phi, ')'))) +
  ggtitle('') +
    facet_grid(age ~ sex) +
  # theme_classic() +
  pres_theme(legend.position = 'top', #panel.border = element_rect(fill = NA),
             plot.title = element_text(hjust = 0.5)) +
  guides(color = guide_legend("", nrow = 1, byrow = T)) +
  scale_color_manual(values = rainbow2[c(2,3,5,6,7,8)], name = '') +
  scale_x_continuous(breaks = c(seq(1, 25, by = 5)), labels = c(seq(2000, (2000+n_occasions-1), by = 5)))


```

### Detection probability
```{r detection}

#mean detection probabilities
p.vals <- post_sum %>%
  filter(!grepl('phi', variable) & grepl('int.p', variable)) %>%
  transform(age = gsub('int.p', '', variable)) %>%
  transform(age = factor(age, levels = c('1', '2', '3', 'J', 'A'),
                         labels = c('1', '2', '3', 'J', 'A')))

ggplot(p.vals, aes(age, med)) +
  geom_point() +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.15) +
  xlab('') + ylab('Detection probability') +
  ggtitle('') +
  plot_theme(legend.position = 'top', panel.border = element_rect(fill = NA),
             plot.title = element_text(hjust = 0.5)) +
  guides(color = guide_legend("", nrow = 1, byrow = T)) +
  scale_color_manual(values = rainbow2[c(2,3,5,6,7,8)], name = '') 

#temporal variance
p.vals <- post_sum %>%
  filter(grepl('.prob', variable)) %>%
  filter(!grepl('phi', variable)) %>%
  transform(year = as.numeric(as.character(gsub('.+\\[([0-9]+)\\].*$', '\\1', 
                                                variable)))) %>%
  transform(variable = sub("\\[.*", "", variable)) %>%
  transform(age = gsub('.prob', '', variable)) %>%
  transform(age = gsub('p', '', age)) %>%
  transform(age = factor(age, levels = c('1', '2', '3', '4', 'A'),
                         labels = c('1', '2', '3', 'J', 'A'))) %>%
  filter(!is.na(age))

ggplot(p.vals, aes(year, med)) +
  geom_errorbar(aes(x = year, ymin=lower, ymax=upper), position = position_dodge(0.5), 
                width = 0.5, show.legend = F) +
  geom_point(size = 0.8, position = position_dodge(0.5)) + 
  geom_line(size = 0.7, position = position_dodge(0.5)) +
  xlab('') + ylab(expression(paste('Detection probability '))) +
  ggtitle('') +
    facet_wrap(. ~ age) +
  # facet_grid(age ~ region) +
  plot_theme(legend.position = 'top', panel.border = element_rect(fill = NA),
             plot.title = element_text(hjust = 0.5)) +
  guides(color = guide_legend("", nrow = 1, byrow = T)) +
  scale_color_manual(values = rainbow2[c(2,3,5,6,7,8)], name = '') +
  scale_x_continuous(breaks = c(seq(1, 19, by = 2)), labels = c(seq(2001, 2019, by = 2)))

```

